---
title: "Descriptive Statistics with R (Practical Part)"
output:
  pdf_document:
    number_sections: yes
  html_document: default
  
---

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
# Load core libraries used throughout this chapter
library(tidyverse)
library(moments)      # skewness, kurtosis
library(data.table)
library(gtsummary)    # Table 1 summaries
library(gt)           # export gtsummary tables
library(skimr)        # quick missingness / distribution overview
library(DescTools)    # BinomCI etc.
library(gridExtra)
library(patchwork)
library(readxl)

# NOTE ON FILE PATHS
# Every path below uses here::here() so the chapter runs on any machine
# once the project's .Rproj / root is set correctly. Replace the folder
# names to match your own project structure if you are following along.

library(here)
data <- haven::read_dta(here("Environment/Datasets", "dt.dta"))
data_new <- read_excel(here("Environment/Datasets", "wcgs.xls"))
```


# Learning Objectives

By the end of this chapter, you should be able to:

1.  Summarize categorical variables using frequencies and proportions.
2.  Summarize numeric variables using measures of central tendency,
    dispersion, and shape (skewness/kurtosis).
3.  Assess whether a numeric variable is approximately normally
    distributed.
4.  Produce cross-tabulations and stratified numeric summaries.
5.  Build a publication-ready "Table 1" using `gtsummary`.
6.  Visualize categorical and numeric data using base R and `ggplot2`.
7.  Produce a simple choropleth map using `sf` and `ggplot2`.



You can get a descriptive overview of the whole dataset using the
`summary()` function.

```{r}
data <- haven::read_dta(here("Environment/Datasets", "dt.dta"))
# clean data
data$Sex <- factor(data$Sex, levels = c(0,1), labels = c("Female", "Male"))

# summarize
summary(data)
```

::: callout-note
Notice that the summary output differs depending on the underlying
data type (numeric vs. factor vs. character).
:::


# Part 1: Descriptive Statistics of Categorical Variables

The categorical variables can be described using frequency and
percentage.


## Frequency

We can use the `table()` function to generate a frequency table for
categorical variables. The `useNA` argument controls whether missing
values are shown: `"no"` (default), `"ifany"`, or `"always"`.
```{r}
table(data$Sex)
table(data$Sex, useNA = "no")
table(data$Sex, useNA = "ifany")  # shows NA column only if missings exist
table(data$Sex, useNA = "always") # always shows NA column
```


## Proportion

To calculate proportions, use `prop.table()`, which is always applied
on top of a `table()` object.

```{r}
sex <- table(data$Sex)
prop.table(sex)

# As percentage, rounded to 1 decimal
round(prop.table(sex) * 100, 1)

# Using the pipe (tidyverse loaded already)
100 * table(data$Sex) |>
  prop.table() |>
  round(3)
```

## Bar Chart and Pie Chart (Base R)

**Pie chart:** `pie(x, labels = names(x), col = NULL, main = NULL, ...)`

```{r}
pie(sex, main = "Sex-wise distribution of population")

pie(sex,
    main = "Sex-wise distribution of population",
    labels = c("Female", "Male"),
    col = c("pink", "lightblue"))
```

**Bar chart:** `barplot(height, names.arg, beside = TRUE, col = NULL, main = NULL, xlab = NULL, ylab = NULL)`

```{r}
barplot(sex,
    main = "Sex-wise distribution of population",
    names.arg = row.names(sex),
    ylab = "Frequency", xlab = "Gender",
    col = c("pink", "lightblue"),
    las = 1)
```

::: callout-tip
Base R plots (`pie()`, `barplot()`, `hist()`, `boxplot()`) are fast and
great for exploratory checks while you're cleaning data. For
publication-quality figures, you'll generally move to `ggplot2` — see
Part 5.
:::

## Cross-Tabulation (Two Categorical Variables)

```{r}
table(data$Sex, data$Ethnicity)  # first variable = rows, second = columns

# dplyr
data |>
  select(Sex, Ethnicity) |>
  table()

# data.table
data.dt <- data.table(data)
data.dt[, table(Sex, Ethnicity)]
```

### Proportions for a Cross-Tabulation

-   default: proportion of the whole table
-   `margin = 1`: row proportion
-   `margin = 2`: column proportion

```{r}
ctable.sex.eth <- table(data$Sex, data$Ethnicity)

prop.table(ctable.sex.eth)             # overall
prop.table(ctable.sex.eth, margin = 1) # row %
prop.table(ctable.sex.eth, margin = 2) # column %

round(prop.table(ctable.sex.eth) * 100, 2)

# data.table equivalent
data.dt[, round(prop.table(table(Sex, Ethnicity), 1) * 100, 1)]
data.dt[, round(prop.table(table(Sex, Ethnicity), 2) * 100, 1)]
```

::: callout-tip
`prop.table()` takes a cross-table produced by `table()` as its first
argument. Only when piping with `|>` can you supply the `margin`
argument unnamed, as shown above.
:::

### Cross-Tabulation Bar Chart

```{r}
barplot(ctable.sex.eth,
    main = "Sex and ethnicity-wise distribution of population",
    ylab = "Frequency", xlab = "Ethnicity",
    beside = TRUE,
    col = c("skyblue", "orange"))
legend("topright", legend = c("Female", "Male"),
       fill = c("skyblue", "orange"))
```

------------------------------------------------------------------------

# Part 2: Descriptive Statistics of Numeric Variables

| Function | Use |
|----|----|
| `mean(x, na.rm=TRUE)` | Arithmetic mean |
| `sd(x, na.rm=TRUE)` | Standard deviation |
| `var(x, na.rm=TRUE)` | Variance |
| `median(x, na.rm=T)` | Median |
| `quantile(x, probs, type)` | Quantiles |
| `min(x, na.rm=T)` / `max(x, na.rm=T)` | Minimum / maximum |
| `range(x, na.rm=T)` | Min and max |
| `moments::skewness(x)` | Skewness |
| `moments::kurtosis(x)` | Kurtosis |

## Missingness Overview First

Before summarizing, it's good habit to check how much data is missing
— this connects back to the missing-data chapter and should always
precede your descriptive statistics.

```{r}
skimr::skim(data)
```

## Central Tendency and Dispersion

```{r}
mean(data$Age, na.rm = TRUE)
sd(data$Age, na.rm = TRUE)
var(data$Age, na.rm = TRUE)
median(data$Age, na.rm = TRUE)
quantile(data$Age, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
min(data$Age, na.rm = TRUE)
max(data$Age, na.rm = TRUE)
range(data$Age, na.rm = TRUE)

# Mode (R has no built-in mode() for this purpose)
names(which.max(table(data$Age)))

# Coefficient of variation — useful when comparing spread
# across variables measured on different scales
cv <- function(x) sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE) * 100
cv(data$Age)
```

::: callout-tip
`mode()` in base R does **not** compute the statistical mode — it
returns the storage type of an object (numeric, integer, list,
factor, etc.). Use the `which.max(table(x))` approach shown above
instead.
:::

## Shape of the Distribution: Skewness and Kurtosis

Mean and median alone can be misleading for skewed variables common in
public health data (length of hospital stay, income, viral load).
Skewness tells you about asymmetry; kurtosis tells you about
"tailedness" relative to a normal distribution.

```{r}
moments::skewness(data_new$age, na.rm = TRUE)
moments::kurtosis(data_new$age, na.rm = TRUE)
```

**Interpreting skewness:**

-   \~0: roughly symmetric
-   \> 0: right-skewed (long tail toward higher values) — mean \>
    median
-   \< 0: left-skewed (long tail toward lower values) — mean \<
    median

**Interpreting kurtosis:** values close to 3 resemble a normal
distribution (some packages report "excess kurtosis" centered at 0 —
always check which convention your function uses).

## Is the Variable Normally Distributed?

Whether a numeric variable is approximately normal informs later
choices (e.g., parametric vs. non-parametric tests). Two common
approaches:

```{r}
#| column: page
#| layout-nrow: 1

# Visual check: Q-Q plot
qqnorm(data_new$age, main = "Q-Q Plot of Age")
qqline(data_new$age, col = "red", lwd = 2)

# Formal test (Shapiro-Wilk; sensitive with large n)
shapiro.test(data_new$age)
```

::: callout-note
With large sample sizes, `shapiro.test()` will often reject normality
even for minor, practically irrelevant deviations. Always pair the
formal test with a visual check (histogram or Q-Q plot) and use
judgment rather than relying on the p-value alone.
:::

## Histogram

`hist(x, main = NULL, xlab = NULL, ylab = NULL, col = NULL, breaks = "Sturges", …)`

```{r}
#| column: page
#| layout-nrow: 1

hist(data$Age,
     main = "Distribution of Age", xlab = "Age", ylab = "Frequency",
     col = "skyblue")

# Overlay a normal curve for visual comparison
sequence <- seq(min(data$Age, na.rm = TRUE), max(data$Age, na.rm = TRUE), 1)
fun <- dnorm(sequence, mean = mean(data$Age, na.rm = TRUE), sd = sd(data$Age, na.rm = TRUE))

hist(data$Age,
     main = "Distribution of Age", xlab = "Age", ylab = "Frequency",
     col = "skyblue", probability = TRUE)
lines(sequence, fun, col = "blue", lwd = 2)
```

## Boxplot

`boxplot(x, notch = FALSE, varwidth = FALSE, col = NULL, names = NULL, …)`

```{r}
boxplot(data$Age, main = "Box plot for Age", col = "skyblue")
points(x = 1, y = mean(data$Age, na.rm = TRUE), col = "red", pch = 16)
```

------------------------------------------------------------------------

# Part 3: Numeric Variables Stratified by a Categorical Variable

## Base R: `tapply()` and `aggregate()`

`tapply(X, INDEX, FUN)` applies a function to subsets of `X` defined by
`INDEX`.

```{r}
tapply(data$Age, data$Sex, mean)
tapply(data$Age, data$Sex, sd)

quantile_fun <- function(x) {
  q <- quantile(x, probs = c(0.25, 0.75))
  names(q) <- c("q1", "q3")
  q
}
tapply(data$Age, data$Sex, quantile_fun)

aggregate(Age ~ Sex, data = data, FUN = mean)
aggregate(Age ~ Sex, data = data, FUN = sd)
```

## dplyr

```{r}
data |>
  group_by(Sex) |>
  summarise(age_mean = mean(Age, na.rm = TRUE),
            age_sd   = sd(Age, na.rm = TRUE),
            age_min  = min(Age, na.rm = TRUE),
            age_q1   = quantile(Age, probs = 0.25, na.rm = TRUE),
            age_q2   = quantile(Age, probs = 0.50, na.rm = TRUE),
            age_q3   = quantile(Age, probs = 0.75, na.rm = TRUE))

data |>
  group_by(Sex, Ethnicity) |>
  summarise(age_mean = mean(Age, na.rm = TRUE),
            age_sd   = sd(Age, na.rm = TRUE), .groups = "drop")
```

## data.table

```{r}
setDT(data)

data[, .(age_mean = mean(Age, na.rm = TRUE),
         age_sd   = sd(Age, na.rm = TRUE),
         age_q1   = quantile(Age, probs = 0.25, na.rm = TRUE),
         age_q3   = quantile(Age, probs = 0.75, na.rm = TRUE)),
     by = Sex]

data[, .(age_mean = mean(Age, na.rm = TRUE),
         age_sd   = sd(Age, na.rm = TRUE)),
     by = .(Sex, Ethnicity)]
```

## Boxplot by Group

```{r}
#| column: page
#| layout-nrow: 1

boxplot(data$Age ~ data$Sex,
        main = "Distribution of Age by Sex",
        col = c("lightpink", "lightblue"))

boxplot(data$Age ~ data$Ethnicity,
        main = "Distribution of Age by Ethnicity",
        col = c("grey", "skyblue", "lightpink", "lightgreen", "lightyellow"))
```

------------------------------------------------------------------------

# Part 4: Building a "Table 1" Summary Table

Nearly every public health manuscript opens with a baseline
characteristics table stratified by exposure or outcome group — often
called "Table 1." The `gtsummary` package builds these in a few lines
and produces output ready to export to Word.

```{r}
table1 <- data |>
  select(Age, Sex, Ethnicity) |>
  tbl_summary(
    by = Sex,
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                      all_categorical() ~ "{n} ({p}%)"),
    missing = "ifany"
  ) |>
  add_p() |>
  add_overall() |>
  modify_header(label ~ "**Variable**") |>
  bold_labels()

table1
```

### Exporting Table 1

```{r, eval=FALSE}
# Export to Word
table1 |>
  as_gt() |>
  gt::gtsave(here("Environment/Outputs", "table1.docx"))

# Export to a plain data frame / Excel
table1_df <- as_tibble(table1)
writexl::write_xlsx(table1_df, here("Environment/Outputs", "table1.xlsx"))
```

------------------------------------------------------------------------

# Part 5: Graphics with ggplot2

Base R plotting (Parts 1–3) is fast for exploratory data analysis.
`ggplot2` is generally the choice for **publication-quality** figures
because of its consistent grammar and fine-grained control over every
visual element — this section rebuilds several of the plots above
using `ggplot2`.

## Choosing a Chart Type

The right chart depends on your variable type and what you want to
communicate:

-   **Nominal** (Sex, Ethnicity, Province): bar chart, pie chart, tree
    map
-   **Ordinal** (education level, Likert scale): ordered bar chart,
    dot plot
-   **Interval/Ratio** (Age, Height, Income): histogram, box plot,
    line chart
-   **Time series**: line chart, area chart
-   **Hierarchical**: treemap, sunburst chart
-   **Correlation/Relationships**: scatter plot, bubble chart, heatmap
-   **Distribution**: histogram, box plot, kernel density plot
-   **Geospatial**: choropleth map, cartogram
-   **Comparison across groups**: grouped or stacked bar chart

## The Grammar of Graphics

Every `ggplot2` plot has three required components:

1.  A **dataset** in tidy format.
2.  **Aesthetic mappings** between variables and visual properties
    (x, y, color, fill, shape, size).
3.  At least one **layer** (`geom_*()`) describing how observations
    are rendered — e.g. `geom_bar()`, `geom_boxplot()`,
    `geom_histogram()`, `geom_point()`, `geom_line()`.

## Basics

```{r}
ggplot(data = data, aes(x = Sex, y = Age)) +
  geom_boxplot()

ggplot(data = data, aes(x = Sex, y = Age, fill = Ethnicity)) +
  geom_boxplot()

ggplot(data = data, aes(x = Sex, y = Age, fill = Ethnicity)) +
  geom_boxplot() +
  xlab("Gender") +
  ylab("Age (in years)") +
  labs(title = "Sex- and ethnicity-wise distribution of Age",
       subtitle = "Caption here", tag = "A")
```

## Manual Colors

```{r}
ggplot(data = data, aes(x = Sex, y = Age, fill = Ethnicity)) +
  geom_boxplot() +
  scale_fill_manual(values = c("lightpink", "lightblue", "lightyellow",
                                "lightgreen", "navy")) +
  xlab("Gender") + ylab("Age (in years)") +
  labs(title = "Sex- and ethnicity-wise distribution of Age")
```

## Faceting

Faceting splits data into subsets and repeats the same plot for each
— `facet_wrap()` (flexible grid) vs `facet_grid()` (strict grid).

```{r}
ggplot(data = data, aes(x = Sex, y = Age, fill = Ethnicity)) +
  geom_boxplot() +
  facet_wrap(facets = ~Ethnicity, nrow = 3, scales = "free")
```

## Themes

`ggplot2` themes are built from four components: theme **elements**
(e.g. `panel.background`, `axis.title`, `legend.title`), **element
functions** (`element_text()`, `element_line()`, `element_rect()`,
`element_blank()`), the `theme()` function to override defaults, and
**complete themes** (`theme_minimal()`, `theme_light()`,
`theme_dark()`, `theme_classic()`).

```{r}
ggplot(data = data, aes(x = Sex, y = Age, fill = Ethnicity)) +
  geom_boxplot() +
  xlab("Gender") + ylab("Age (in years)") +
  labs(title = "Sex- and ethnicity-wise distribution of Age") +
  facet_wrap(facets = ~Ethnicity, nrow = 3, scales = "free_x") +
  theme(panel.background = element_rect(fill = "lightgreen"),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold", color = "red", size = 13),
        axis.title = element_text(face = "bold", color = "blue"),
        legend.position = "bottom")
```

------------------------------------------------------------------------

# Part 6: Mapping with sf and ggplot2

```{r}
llmap <- sf::st_read(here("Environment/Datasets", "Shape file of Nepal", "Nepal_llmapcoded.shp"))

ggplot(data = llmap) +
  geom_sf(aes(geometry = geometry), col = "black") +
  coord_sf(crs = "WGS84") +
  theme(panel.background = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

ggplot(data = llmap) +
  geom_sf(aes(geometry = geometry, fill = TYPE), col = "black") +
  scale_fill_viridis_d() +
  coord_sf(crs = "WGS84") +
  theme(panel.background = element_blank(),
        legend.title = element_text(face = "bold"),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())
```

------------------------------------------------------------------------

# Part 7: Applied Case Studies

The examples in this section use different datasets from the rest of
the chapter (introduced individually below) to show how the tools
above come together in real analysis workflows.

## Case Study 1: 4+ ANC Visits by Province, with Confidence Intervals

**Dataset:** `For graphics1.RDS` — survey data with variables
`Province`, `Location`, and `Four_ANC` (whether a woman had 4+
antenatal care visits).

```{r}
dt <- readRDS(here("Environment/Datasets", "For graphics1.RDS"))

a <- dt |>
  group_by(Location, Province, Four_ANC) |>
  tally() |>
  ungroup() |>
  group_by(Location, Province) |>
  mutate(total = sum(n),
         percent = n / total * 100,
         ci = DescTools::BinomCI(n, total, method = "clopper-pearson") * 100) |>
  filter(Four_ANC == "Yes")

ggplot(data = a) +
  geom_bar(aes(x = Location, y = percent, fill = Location),
           col = "black", stat = "identity", alpha = 1) +
  geom_errorbar(aes(x = Location, y = percent,
                     ymin = a$ci[, 2], ymax = a$ci[, 3]),
                width = 0.3, linewidth = 1) +
  facet_wrap(~Province, nrow = 1) +
  geom_text(aes(x = Location, y = percent, label = sprintf("%0.2f", percent)),
            size = 4, vjust = 1.5, hjust = -0.1, angle = 90, col = "blue") +
  scale_fill_manual(values = c("pink", "lightgreen")) +
  xlab("Location") + ylab("Percent (%) with CI") +
  labs(title = "Distribution of 4+ ANC by Location and Province") +
  theme(panel.background = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom")
```

::: callout-note
Notice how the percent was calculated *within* `Location` and
`Province` groups before faceting — a common pivot-table mistake is
computing percentages over the wrong grouping level and only noticing
once the bars don't sum to 100% within each panel.
:::

## Case Study 2: Global Burden of Disease — Diabetes Mellitus

**Dataset:** `For graphics2.xlsx` (sheet `"dataset"`) — GBD-style data
with `Age`, `Sex`, `Class` (DM type), and `prop` (prevalence/death/DALY
proportions).

```{r}
gbd <- read_excel(here("Environment/Datasets", "For graphics2.xlsx"), sheet = "dataset")

gbd$Age2 <- factor(gbd$Age,
  levels = c("<1","1-4","5-9","10-14","15-19","20-24","25-29","30-34",
             "35-39","40-44","45-49","50-54","55-59","60-64","65-69","70+"),
  ordered = TRUE)

gbd$Class <- factor(gbd$Class, levels = c("DM", "DM1", "DM2"),
  labels = c("Diabetes Mellitus (combined)",
             "Diabetes Mellitus (Type 1)",
             "Diabetes Mellitus (Type 2)"))

prevalence <- gbd[gbd$Ovearll_class %in% c("DM_P", "DM1_P", "DM2_P"), ]
death      <- gbd[gbd$Ovearll_class %in% c("DM_D", "DM1_D", "DM2_D"), ]
daly       <- gbd[gbd$Ovearll_class %in% c("DM_Da", "DM1_Da", "DM2_Da"), ]

c1 <- ggplot(prevalence) +
  geom_bar(aes(x = Age2, y = prop, fill = Sex), stat = "identity", position = "dodge") +
  xlab("Age (years)") + ylab("Prevalence") +
  facet_wrap(~Class, nrow = 3, scales = "free_x") +
  theme(legend.position = "none", strip.text = element_blank()) +
  coord_flip()

c2 <- ggplot(death) +
  geom_bar(aes(x = Age2, y = prop, fill = Sex), stat = "identity", position = "dodge") +
  xlab("") + ylab("Deaths") +
  facet_wrap(~Class, nrow = 3, scales = "free_x") +
  theme(legend.position = "none", strip.text = element_blank()) +
  coord_flip()

c3 <- ggplot(daly) +
  geom_bar(aes(x = Age2, y = prop, fill = Sex), stat = "identity", position = "dodge") +
  xlab("") + ylab("DALYs") +
  facet_wrap(~Class, nrow = 3, scales = "free_x", strip.position = "right") +
  theme(legend.position = "none") +
  coord_flip()

combined <- (c1 + c2 + c3) & theme(legend.position = "top", legend.title = element_blank())
m <- combined + plot_layout(guides = "collect")
m
```

```{r, eval=FALSE}
ggsave(here("Environment/Outputs", "chart1.pdf"), plot = m, dpi = 300)
```

## Case Study 3: GBD Trends Over Time — Chronic Kidney Disease

**Dataset:** `For graphics3.xlsx` — raw GBD export with `measure_name`,
`age_name`, `sex_name`, `cause_name`, `year`, `val`.

```{r}
gbd3 <- read_excel(here("Environment/Datasets", "For graphics3.xlsx"))

gbd3$Year <- factor(gbd3$year, ordered = TRUE)
gbd3$cause_name <- as.factor(gbd3$cause_name)
gbd3$sex_name <- as.factor(gbd3$sex_name)

daly3 <- gbd3[gbd3$measure_name == "DALYs (Disability-Adjusted Life Years)", ]
deaths3 <- gbd3[gbd3$measure_name == "Deaths", ]
prev3 <- gbd3[gbd3$measure_name == "Prevalence", ]

pe <- prev3 |>
  filter(sex_name == "Both", metric_name == "Rate",
         !age_name %in% c("Age-standardized", "All ages", "70+"),
         cause_name != "Chronic kidney disease",
         year %in% c(1990, 2019))

a <- ggplot(pe) +
  geom_bar(aes(x = age_name, y = val, fill = cause_name),
           stat = "identity", position = "stack", color = "black", linewidth = 0.01) +
  ylab("Prevalence rate (per 100,000)") + xlab("") +
  scale_fill_manual(values = c("#9CD5E8", "#E69F00", "#00559C", "#CD534C", "#6BB43E")) +
  facet_wrap(~Year) +
  theme(legend.position = "bottom", legend.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

a
```

```{r, eval=FALSE}
ggsave(here("Environment/Outputs", "chart2.png"), plot = a, dpi = 300)
```

------------------------------------------------------------------------

# Exercises

1.  Using `dt.dta`, produce a frequency table and bar chart for a
    categorical variable other than `Sex`.
2.  Compute skewness and kurtosis for `Age`. Is the distribution
    symmetric? Support your answer with a histogram.
3.  Run `shapiro.test()` on `Age` stratified by `Sex` (hint:
    `tapply()`). Do both groups appear normally distributed?
4.  Build a `gtsummary` Table 1 stratified by `Ethnicity` instead of
    `Sex`, and export it to Word.
5.  Recreate the Case Study 1 bar chart using a different confidence
    interval method available in `DescTools::BinomCI()` (e.g.
    `"wilson"`) and compare the resulting intervals.
