---
title: "Discrete Probability Distributions"
output:
  pdf_document:
    number_sections: yes
  html_document: default
editor_options:
  chunk_output_type: console
editor: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
library(ggplot2)
```

## Random Variables

A **random variable** is a numeric variable whose value is determined by the outcome of a random experiment.
Random variables are denoted with capital letters ($X$, $Y$) and their specific realized values with lowercase letters ($x$, $y$).

-   A **discrete random variable** takes a countable number of values (finite or countably infinite), e.g., number of ER visits in a month.
-   A **continuous random variable** takes any value within an interval (covered in Section 8), e.g., blood pressure.

## Probability Mass Function (PMF)

For a discrete random variable $X$, the **probability mass function (PMF)**, $P(X=x)$, gives the probability that $X$ takes on each specific value $x$.
A valid PMF must satisfy:

1.  $0 \le P(X=x) \le 1$ for every $x$
2.  $\sum_{\text{all } x} P(X=x) = 1$

The **cumulative distribution function (CDF)** is $F(x) = P(X \le x) = \sum_{k\le x} P(X=k)$.

## Expected Value

The **expected value** (mean) of a discrete random variable is the long-run average value, weighted by probability: $$E(X) = \mu = \sum_{\text{all } x} x\, P(X=x)$$

## Variance

$$Var(X) = \sigma^2 = \sum_{\text{all }x}(x-\mu)^2 P(X=x) = E(X^2) - [E(X)]^2$$

**Worked example:** A community health worker screens patients.
let $X$ = number of positive screens out of the first 3 patients tested, with PMF:

$P(0)=0.20$

$P(1)=0.35$

$P(2)=0.30$

$P(3)=0.15$

**Step 1:** Verify PMF sums to 1:

$0.20+0.35+0.30+0.15=1.00$

**Step 2:** $E(X) = 0(0.20)+1(0.35)+2(0.30)+3(0.15) = 0+0.35+0.60+0.45=1.40$

**Step 3:**$E(X^2) = 0^2(0.20)+1^2(0.35)+2^2(0.30)+3^2(0.15)=0+0.35+1.20+1.35=2.90$

**Step 4:**

$Var(X) = 2.90-1.40^2=2.90-1.96=0.94$

$SD(X)=\sqrt{0.94}=0.97$.

------------------------------------------------------------------------

## Bernoulli Distribution

### Definition

Models a **single trial** with exactly two possible outcomes: "success" (1) or "failure" (0).

### PMF

$$P(X=x) = p^x(1-p)^{1-x}, \quad x \in \{0,1\}$$

### Parameters, Mean, Variance

| Parameter | $p$ = probability of success |
|-----------|------------------------------|
| Mean      | $E(X) = p$                   |
| Variance  | $Var(X) = p(1-p)$            |

**Derivation of mean:**

$E(X) = 0\cdot(1-p) + 1\cdot p = p$

**Derivation of variance:**

$E(X^2) = 0^2(1-p)+1^2(p) = p$

$Var(X) = E(X^2)-[E(X)]^2 = p-p^2=p(1-p)$.

### Shape & Assumptions

-   Two-point distribution

-   Shape depends entirely on $p$.

-   Assumes a single trial with a fixed success probability.

### Applications / Public Health Example

Whether a single patient tests positive for a disease ($p$ = prevalence); whether a single newborn is preterm; the building block for the Binomial distribution.

### R Code and Graph

```{r bernoulli}
p <- 0.3
x <- c(0, 1)
prob <- dbinom(x, size = 1, prob = p)
ggplot(data.frame(x = factor(x), prob), aes(x, prob)) +
  geom_col(fill = "steelblue") +
  labs(title = "Bernoulli(p = 0.3)", x = "X", y = "P(X = x)") +
  theme_minimal()
```

### Manual Calculation

If prevalence of a risk factor 30% ($p = 0.30$).

$P(X=1) = 0.30$

$P(X=0)=0.70$

Mean $=0.30$

Variance $=0.30(0.70)=0.21$

------------------------------------------------------------------------

## Binomial Distribution

### Definition

Models the number of **successes in** $n$ independent Bernoulli trials, each with the same probability of success $p$.

### PMF

$$P(X=x) = \binom{n}{x}p^x(1-p)^{n-x}, \quad x=0,1,\ldots,n$$

where $\binom{n}{x} = \dfrac{n!}{x!(n-x)!}$.

### Parameters, Mean, Variance

| Parameter | $n$ = number of trials, $p$ = probability of success |
|-----------|------------------------------------------------------|
| Mean      | $E(X) = np$                                          |
| Variance  | $Var(X) = np(1-p)$                                   |

### Shape

-   Symmetric when $p=0.5$

-   Right-skewed when $p<0.5$ and left-skewed when $p>0.5$

-   Approaches a Normal shape as $n$ grows (for fixed $p$ not too close to 0 or 1).
    It is a consequence of the Central Limit Theorem.

### Assumptions

1.  Fixed number of trials $n$.
2.  Each trial has only two outcomes.
3.  Constant probability of success $p$ across trials.
4.  Trials are independent.

### Applications / Public Health Example

Number of patients who respond to treatment out of $n$ treated; number of children vaccinated out of $n$ eligible; number of surgical-site infections out of $n$ surgeries.

### R Code and Graph

```{r binomial}
n <- 20; p <- 0.3
x <- 0:n
prob <- dbinom(x, size = n, prob = p)
ggplot(data.frame(x, prob), aes(x, prob)) +
  geom_col(fill = "darkorange") +
  labs(title = "Binomial(n = 20, p = 0.3)", x = "Number of successes", y = "P(X = x)") +
  theme_minimal()
```

### Worked Example (Manual Calculation)

In a clinic, 30% of screened patients ($p=0.3$) are found to have elevated cholesterol.
Out of $n=5$ patients screened today, what is the probability exactly 2 have elevated cholesterol?

**Step 1:** Identify $n=5$, $x=2$, $p=0.3$

**Step 2:** Compute the binomial coefficient: $\binom{5}{2} = \dfrac{5!}{2!3!} = \dfrac{120}{2\times 6}=10$

**Step 3:** Compute $p^x = 0.3^2 = 0.09$ and $(1-p)^{n-x} = 0.7^3=0.343$.

**Step 4:** Multiply: $$P(X=2) = 10 \times 0.09 \times 0.343 = 0.3087 \approx 30.9\%$$

**Verification in R:** `dbinom(2, size = 5, prob = 0.3)` returns 0.3087.

Mean $= np = 5(0.3) = 1.5$ patients

Variance $= np(1-p)=5(0.3)(0.7)=1.05$

------------------------------------------------------------------------

## Geometric Distribution

### Definition

Models the **number of trials needed to obtain the first success** in a sequence of independent Bernoulli trials.

### PMF

$$P(X=x) = (1-p)^{x-1}p, \quad x=1,2,3,\ldots$$

(Convention: $X$ counts the trial *on which* the first success occurs. An alternative parameterization counts the number of *failures before* the first success, $Y=X-1$.)

### Parameters, Mean, Variance

| Parameter | $p$ = probability of success per trial |
|-----------|----------------------------------------|
| Mean      | $E(X) = \dfrac{1}{p}$                  |
| Variance  | $Var(X) = \dfrac{1-p}{p^2}$            |

### Shape & Assumptions

-   Always right-skewed

-   Highest probability at $x=1$, decreasing geometrically thereafter.

-   Assumes independent trials with constant $p$, continuing until the first success (a "waiting time" distribution).

### Applications / Public Health Example

-   Number of couples' pregnancy attempts (cycles) until conception

<!-- -->

-   Number of blood donors screened until finding one with a specific rare blood type

-   Number of contact-tracing attempts until successfully reaching a contact.

### R Code and Graph

```{r geometric}
p <- 0.25
x <- 1:15
prob <- dgeom(x - 1, prob = p)  # R's dgeom counts failures before success
ggplot(data.frame(x, prob), aes(x, prob)) +
  geom_col(fill = "seagreen") +
  labs(title = "Geometric(p = 0.25), X = trial of first success",
       x = "Trial number of first success", y = "P(X = x)") +
  theme_minimal()
```

::: callout-tip
R's `dgeom(k, prob)` defines $k$ as the number of **failures before** the first success, so to match the "trial number" parameterization above, use `dgeom(x - 1, prob)`.
:::

### Worked Example (Manual Calculation)

If 25% of blood donors ($p=0.25$) have blood type O-negative, what is the probability that the **4th** donor screened is the first one with O-negative blood?

**Step 1:**

$x=4$, $p=0.25$, so $1-p=0.75$

**Step 2:** Apply the PMF\
$$P(X=4) = (0.75)^{3}(0.25) = 0.4219\times0.25 = 0.1055 \approx 10.6\%$$

**Mean** $= 1/0.25 = 4$ donors (on average, expect to screen 4 donors to find one O-negative).

**Variance** $= (1-0.25)/0.25^2 = 0.75/0.0625=12$.

------------------------------------------------------------------------

## Negative Binomial Distribution

### Definition

Generalizes the Geometric distribution: models the **number of trials needed to achieve the** $r$-th success.

### PMF

$$P(X=x) = \binom{x-1}{r-1}p^r(1-p)^{x-r}, \quad x=r,r+1,r+2,\ldots$$

### Parameters, Mean, Variance

| Parameter | $r$ = target number of successes, $p$ = probability of success per trial |
|------------------------|-----------------------------------------------|
| Mean | $E(X) = \dfrac{r}{p}$ |
| Variance | $Var(X) = \dfrac{r(1-p)}{p^2}$ |

### Shape & Assumptions

-   Right-skewed (less so as $r$ increases)

-   Same independence/constant-$p$ assumptions as the Geometric distribution (which is the special case $r=1$).

-   Also widely used (in its alternative "count" parameterization) to model **overdispersed count data**, count data whose variance exceeds its mean, which the Poisson distribution cannot accommodate.

### Applications / Public Health Example

-   Number of patients that must be screened to identify $r=3$ cases of a rare disease for a case series.

-   Modeling overdispersed disease counts (e.g., hospital readmission counts) where a Poisson model underestimates variability.

### R Code and Graph

```{r negbinom}
r <- 3; p <- 0.25
x <- r:30
prob <- dnbinom(x - r, size = r, prob = p)  # R counts failures before r-th success
ggplot(data.frame(x, prob), aes(x, prob)) +
  geom_col(fill = "purple") +
  labs(title = "Negative Binomial(r = 3, p = 0.25)",
       x = "Trial of 3rd success", y = "P(X = x)") +
  theme_minimal()
```

### Worked Example (Manual Calculation)

Continuing the O-negative blood donor example ($p=0.25$): what is the probability that the **6th** donor screened is the one who provides the **3rd** O-negative unit found ($r=3$)?

**Step 1:**

$x=6$, $r=3$, $p=0.25$

**Step 2:**

Binomial coefficient: $\binom{x-1}{r-1} = \binom{5}{2} = 10$

**Step 3:**

$p^r = 0.25^3 = 0.015625$

$(1-p)^{x-r} = 0.75^3 = 0.421875$

**Step 4:** $$P(X=6) = 10\times 0.015625\times 0.421875 = 0.0659 \approx 6.6\%$$

Mean $= r/p = 3/0.25=12$ donors expected to find 3 O-negative units.

------------------------------------------------------------------------

## Hypergeometric Distribution

### Definition

Models the number of successes in $n$ draws **without replacement** from a finite population of size $N$ containing $K$ successes, used when sampling is done **without replacement** from a small, finite population (in contrast to the Binomial, which assumes independence/replacement).

### PMF

$$P(X=x) = \frac{\binom{K}{x}\binom{N-K}{n-x}}{\binom{N}{n}}$$

### Parameters, Mean, Variance

| Parameter | $N$ = population size, $K$ = number of successes in population, $n$ = sample size |
|--------------------------|----------------------------------------------|
| Mean | $E(X) = n\dfrac{K}{N}$ |
| Variance | $Var(X) = n\dfrac{K}{N}\cdot\dfrac{N-K}{N}\cdot\dfrac{N-n}{N-1}$ |

The variance formula includes the **finite population correction factor**, $\dfrac{N-n}{N-1}$, which distinguishes it from the Binomial.

### Shape & Assumptions

-   Similar in shape to the Binomial for large $N$ relative to $n$ (in fact, the Hypergeometric converges to the Binomial as $N\to\infty$ with $K/N=p$ fixed).

-   Assumes sampling **without replacement** from a finite, known population.

### Applications / Public Health Example

Selecting a sample of vials for quality control from a finite batch (as in Section 4's defective-vial example); selecting a committee/sample from a small finite roster of patients with known disease status; capture-recapture population size estimation.

### R Code and Graph

```{r hypergeom}
N <- 50; K <- 10; n <- 8
x <- 0:min(n, K)
prob <- dhyper(x, m = K, n = N - K, k = n)
ggplot(data.frame(x, prob), aes(x, prob)) +
  geom_col(fill = "firebrick") +
  labs(title = "Hypergeometric(N=50, K=10, n=8)", x = "Number of successes", y = "P(X = x)") +
  theme_minimal()
```

### Worked Example (Manual Calculation)

A batch of $N=50$ vaccine vials contains $K=10$ expired vials.
A quality inspector randomly selects $n=5$ vials for testing (without replacement).
What is the probability exactly $x=2$ are expired?

**Step 1:**

$\binom{K}{x} = \binom{10}{2} = 45$

**Step 2:**

$\binom{N-K}{n-x} = \binom{40}{3} = 9880$

**Step 3:**

$\binom{N}{n} = \binom{50}{5} = 2{,}118{,}760$

**Step 4:** $$P(X=2) = \frac{45\times9880}{2{,}118{,}760} = \frac{444{,}600}{2{,}118{,}760}=0.2099\approx21.0\%$$

Mean $= n(K/N) = 5(10/50)=1.0$ expired vial expected in the sample.

------------------------------------------------------------------------

## Poisson Distribution

### Definition

Models the number of **rare, independent events occurring in a fixed interval** of time, space, or exposure, when events happen at a constant average rate $\lambda$.

### PMF

$$P(X=x) = \frac{e^{-\lambda}\lambda^x}{x!}, \quad x=0,1,2,\ldots$$

### Parameters, Mean, Variance

| Parameter | $\lambda$ = mean (and rate) of events per interval |
|-----------|----------------------------------------------------|
| Mean      | $E(X) = \lambda$                                   |
| Variance  | $Var(X) = \lambda$                                 |

The Poisson is unique in that its **mean equals its variance**, an assumption that is often violated in real disease-count data (overdispersion), motivating the use of the Negative Binomial as an alternative.

### Shape & Assumptions

-   Right-skewed for small $\lambda$, approaching a symmetric, Normal-like shape as $\lambda$ increases.

-   Assumes:

    -   events occur independently,

    -   at a constant average rate,

    -   two events cannot occur at exactly the same instant,

    -   the number of events in non-overlapping intervals is independent.

### Applications / Public Health Example

Number of new cases of a rare disease reported per month in a district; number of hospital-acquired infections per 1000 patient-days; number of deaths from a rare adverse drug reaction; number of disease outbreaks per year in a region.
The Poisson also arises as the limiting distribution of the Binomial when $n$ is large, $p$ is small, and $np=\lambda$ is moderate (the "law of rare events").

### R Code and Graph

```{r poisson}
lambda <- 4
x <- 0:15
prob <- dpois(x, lambda = lambda)
ggplot(data.frame(x, prob), aes(x, prob)) +
  geom_col(fill = "slateblue") +
  labs(title = "Poisson(lambda = 4)", x = "Number of events", y = "P(X = x)") +
  theme_minimal()
```

### Worked Example (Manual Calculation)

A district reports an average of $\lambda=4$ cases of a rare disease per month.
What is the probability that exactly 6 cases are reported next month?

**Step 1:**

$\lambda=4$, $x=6$

**Step 2:**

$e^{-\lambda} = e^{-4} = 0.01832$

**Step 3:**

$\lambda^x = 4^6=4096$

**Step 4:**

$x! = 6! = 720$

**Step 5:** $$P(X=6) = \frac{0.01832\times4096}{720} = \frac{75.03}{720}=0.1042\approx10.4\%$$

**Verification in R:** `dpois(6, lambda = 4)` returns 0.1042.

**Probability of zero cases:** $P(X=0) = e^{-4}(4^0)/0! = e^{-4} = 0.0183$.

It means 1.8% chance of no cases reported at all next month.

------------------------------------------------------------------------

## Comparison Table — Discrete Distributions

| Distribution | Models | Parameters | Mean | Variance | Trials fixed? | Sampling |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| Bernoulli | Single success/failure | $p$ | $p$ | $p(1-p)$ | 1 trial |  |
| Binomial | \# successes in $n$ trials | $n,p$ | $np$ | $np(1-p)$ | Yes | With replacement / independent |
| Geometric | \# trials to 1st success | $p$ | $1/p$ | $(1-p)/p^2$ | No (random) | With replacement / independent |
| Negative Binomial | \# trials to $r$-th success | $r,p$ | $r/p$ | $r(1-p)/p^2$ | No (random) | With replacement / independent |
| Hypergeometric | \# successes in $n$ draws, finite pop. | $N,K,n$ | $nK/N$ | $n\frac{K}{N}\frac{N-K}{N}\frac{N-n}{N-1}$ | Yes | **Without** replacement |
| Poisson | \# events in fixed interval | $\lambda$ | $\lambda$ | $\lambda$ | N/A (rate-based) |  |

## Decision Guide: Which Distribution Should I Use?

```{mermaid}
flowchart TD
    Q1{"Are you counting events<br/>per unit time/space<br/>(no fixed 'n')?"} -->|Yes| Poisson["Use Poisson"]
    Q1 -->|No, fixed number of trials/draws| Q2{"Sampling with or without<br/>replacement from a<br/>finite, small population?"}
    Q2 -->|Without replacement, finite pop.| Hyper["Use Hypergeometric"]
    Q2 -->|With replacement / independent trials, or infinite population| Q3{"Are you counting<br/>successes in n fixed trials,<br/>or waiting for success(es)?"}
    Q3 -->|"Counting successes in fixed n"| Q4{"n = 1 trial?"}
    Q4 -->|Yes| Bern["Use Bernoulli"]
    Q4 -->|No, n > 1| Binom["Use Binomial"]
    Q3 -->|"Waiting for success(es)"| Q5{"Waiting for 1st success<br/>or r-th success?"}
    Q5 -->|"1st success"| Geo["Use Geometric"]
    Q5 -->|"r-th success"| NB["Use Negative Binomial"]
```

## Summary

-   A **random variable** maps outcomes of a random experiment to numbers; its **PMF** assigns probabilities to each possible value.
-   **Expected value** and **variance** summarize the center and spread of a discrete distribution.
-   **Bernoulli** (1 trial) generalizes to **Binomial** ($n$ trials); **Geometric** (wait for 1st success) generalizes to **Negative Binomial** (wait for $r$-th success).
-   **Hypergeometric** is the without-replacement analogue of the Binomial for finite populations.
-   **Poisson** models rare events over a fixed interval and is the limiting case of the Binomial as $n\to\infty$, $p\to0$, with $np=\lambda$ fixed.

Next section will deal with the probability distribution of **continuous variables**.
