---
title: "Continuous Probability Distributions"
output:
  pdf_document:
    number_sections: yes
  html_document: default
editor_options:
  chunk_output_type: console
editor: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
library(ggplot2)
```

## Probability Density Function (PDF)

For a **continuous random variable**, probability is not assigned to individual points (which each have probability zero) but to **intervals**, via the **probability density function (PDF)**, $f(x)$.
The probability that $X$ falls between $a$ and $b$ is the area under the curve: $$P(a \le X \le b) = \int_a^b f(x)\,dx$$

A valid PDF satisfies $f(x) \ge 0$ for all $x$, and $\int_{-\infty}^{\infty} f(x)\,dx = 1$ (total area under the curve equals 1).

## Cumulative Distribution Function (CDF)

$$F(x) = P(X \le x) = \int_{-\infty}^{x} f(t)\,dt$$

$F(x)$ gives the cumulative probability up to $x$, and is used to find percentiles and probabilities directly.

## Expected Value and Variance

$$E(X) = \mu = \int_{-\infty}^{\infty} x f(x)\,dx \qquad Var(X) = \sigma^2 = \int_{-\infty}^{\infty}(x-\mu)^2 f(x)\,dx$$

------------------------------------------------------------------------

## Uniform Distribution

### Definition

All values in an interval $[a,b]$ are **equally likely**.

### PDF and CDF

$$f(x) = \frac{1}{b-a}, \quad a \le x \le b \qquad $$ $$F(x) = \frac{x-a}{b-a}$$

### Parameters, Mean, Variance

| Parameter | $a$ = minimum, $b$ = maximum |
|-----------|------------------------------|
| Mean      | $\dfrac{a+b}{2}$             |
| Variance  | $\dfrac{(b-a)^2}{12}$        |

### Properties

Constant (flat) density; symmetric about the midpoint $(a+b)/2$.

### Applications / Public Health Example

Modeling a patient's arrival time within a clinic's operating hours as "equally likely" across the interval; simulation studies (uniform random number generation is the basis for simulating other distributions).

### Worked Example

Appointment arrival times are uniformly distributed between 8:00 AM ($a=0$) and 5:00 PM ($b=9$ hours).
What is the probability a patient arrives in the first hour?
$$P(0\le X\le1) = \frac{1-0}{9-0} = 0.111 \approx 11.1\%$$

### R Code and Graph

```{r uniform}
ggplot(data.frame(x = c(0, 9)), aes(x)) +
  stat_function(fun = dunif, args = list(min = 0, max = 9), geom = "area", fill = "skyblue", alpha = 0.5) +
  labs(title = "Uniform(0, 9)", x = "Hours after 8:00 AM", y = "Density") + theme_minimal()
```

------------------------------------------------------------------------

## Normal (Gaussian) Distribution

### Definition

The most important distribution in statistics is a normal or Gaussian distribution.
It is a symmetric, bell-shaped distribution characterized fully by its mean and standard deviation.
Many biological measurements (height, blood pressure, IQ) are approximately Normal, and the Normal distribution is central to the Central Limit Theorem underlying most inferential methods (Sections 9-11).

### PDF

$$f(x) = \frac{1}{\sigma\sqrt{2\pi}}\exp\left(-\frac{(x-\mu)^2}{2\sigma^2}\right), \quad -\infty<x<\infty$$

### Parameters, Mean, Variance

| Parameter | $\mu$ = mean, $\sigma$ = standard deviation |
|-----------|---------------------------------------------|
| Mean      | $\mu$                                       |
| Variance  | $\sigma^2$                                  |

### Properties

-   Symmetric, bell-shaped, unimodal

-   Mean = Median = Mode.

-   Defined by the empirical rule:

    -   \~68% of values fall within $\mu\pm1\sigma$

    -   \~95% within $\mu\pm2\sigma$

    -   \~99.7% within $\mu\pm3\sigma$

-   Notation: $X\sim N(\mu,\sigma^2)$.

### Applications / Public Health Example

Birth weight, height, blood pressure, cholesterol levels are frequently modeled as approximately Normal; the sampling distribution of the sample mean is approximately Normal for large $n$ (Central Limit Theorem), justifying most confidence intervals and hypothesis tests in Sections 7.

### R Code and Graph

```{r normal}
ggplot(data.frame(x = c(40, 160)), aes(x)) +
  stat_function(fun = dnorm, args = list(mean = 100, sd = 15), geom = "area", fill = "tomato", alpha = 0.5) +
  labs(title = "Normal(mean = 100, SD = 15)", x = "X", y = "Density") + theme_minimal()
```

------------------------------------------------------------------------

## Standard Normal Distribution

### Definition

The special case of the Normal distribution with $\mu=0$ and $\sigma=1$, denoted $Z \sim N(0,1)$.
Any Normal variable can be converted ("standardized") to the standard Normal using the **z-score**: $$Zscore = \frac{X-\mu}{\sigma}$$

### PDF

$$f(z) = \frac{1}{\sqrt{2\pi}}e^{-z^2/2}$$

### Worked Example (Manual Calculation)

Adult male height is Normally distributed with $\mu=170$ cm, $\sigma=7$ cm.
What proportion of men are taller than 180 cm?

**Step 1:** Compute the z-score: $z = \dfrac{180-170}{7} = 1.43$

**Step 2:** Look up (or compute via software) $P(Z>1.43)$: from the standard Normal table, $P(Z\le1.43)=0.9236$, so $P(Z>1.43) = 1-0.9236=0.0764$

**Step 3:** Interpretation: approximately **7.6%** of adult men are taller than 180 cm.

**In R:** `1 - pnorm(180, mean = 170, sd = 7)` returns 0.0764.

### R Code and Graph

```{r stdnormal}
ggplot(data.frame(x = c(-4, 4)), aes(x)) +
  stat_function(fun = dnorm, geom = "area", fill = "gray70", alpha = 0.5,
                xlim = c(1.43, 4)) +
  stat_function(fun = dnorm) +
  labs(title = "Standard Normal Z ~ N(0,1); shaded area = P(Z > 1.43)",
       x = "z", y = "Density") + theme_minimal()
```

------------------------------------------------------------------------

## Exponential Distribution

### Definition

Models the **waiting time (or distance)** until the next event in a process where events occur at a constant average rate — the continuous analogue of the Geometric distribution, closely related to the Poisson process.

### PDF and CDF

$$f(x) = \lambda e^{-\lambda x}, \quad x\ge0 \qquad F(x) = 1-e^{-\lambda x}$$

### Parameters, Mean, Variance

| Parameter | $\lambda$ = rate parameter |
|-----------|----------------------------|
| Mean      | $1/\lambda$                |
| Variance  | $1/\lambda^2$              |

### Properties

-   Right-skewed;

-   Exhibits the **memoryless property**: $P(X>s+t\mid X>s)=P(X>t)$

    \[The remaining waiting time does not depend on how long you have already waited\]

### Applications / Public Health Example

Time between successive patient arrivals at an emergency department; survival time in simple constant-hazard survival models; time until equipment failure in a medical device.

### Worked Example

Patients arrive at an ED at an average rate of $\lambda=3$ per hour.
What is the probability the next patient arrives within 10 minutes (1/6 hour)?

**Step 1:** $F(x) = 1-e^{-\lambda x}$, with $\lambda=3$, $x=1/6$

**Step 2:** $\lambda x = 3\times(1/6) = 0.5$

**Step 3:** $$P(X\le 1/6) = 1-e^{-0.5} = 1-0.6065=0.3935\approx39.4\%$$

### R Code and Graph

```{r exponential}
ggplot(data.frame(x = c(0, 2)), aes(x)) +
  stat_function(fun = dexp, args = list(rate = 3), geom = "area", fill = "darkgreen", alpha = 0.5) +
  labs(title = "Exponential(rate = 3 per hour)", x = "Hours until next arrival", y = "Density") + theme_minimal()
```

------------------------------------------------------------------------

## Gamma Distribution

### Definition

Generalizes the Exponential distribution to model the **waiting time until the** $k$-th event in a Poisson process (analogous to how Negative Binomial generalizes Geometric in the discrete case).

### PDF

$$f(x) = \frac{\lambda^k x^{k-1}e^{-\lambda x}}{\Gamma(k)}, \quad x>0$$ where $\Gamma(k)$ is the Gamma function (for integer $k$, $\Gamma(k)=(k-1)!$).

### Parameters, Mean, Variance

| Parameter | $k$ (shape), $\lambda$ (rate) |
|-----------|-------------------------------|
| Mean      | $k/\lambda$                   |
| Variance  | $k/\lambda^2$                 |

### Properties

-   Right-skewed for small $k$, approaching Normal shape as $k$ increases

-   Exponential distribution is the special case at $k=1$

-   Chi-square distribution (below) is also a special case of the Gamma.

### Applications / Public Health Example

Modeling time until $k$ disease relapses occur; modeling right-skewed cost or length-of-stay data in health economics; modeling total rainfall or time until $k$-th case in an outbreak.

### R Code and Graph

```{r gamma}
ggplot(data.frame(x = c(0, 10)), aes(x)) +
  stat_function(fun = dgamma, args = list(shape = 3, rate = 1), geom = "area", fill = "orchid", alpha = 0.5) +
  labs(title = "Gamma(shape = 3, rate = 1)", x = "X", y = "Density") + theme_minimal()
```

::: callout-note
**A Common Thread: Degrees of Freedom**

Starting with the Chi-square distribution below, a new kind of parameter appears: **degrees of freedom** (df, often denoted $k$, $\nu$, or $d$).
Unlike $\mu$ and $\sigma$ for the Normal distribution, degrees of freedom don't describe a population directly - they describe how much *independent information* went into building the statistic whose distribution we're describing.

Each of the next three distributions uses df in a slightly different way, because each arises from a different kind of calculation:

-   **Chi-square:** $k$ independent standard Normal variables are squared and summed, df simply counts how many independent squared terms went into the sum.
-   **Student's t:** built from a sample of size $n$, where one piece of information is used up estimating the sample mean before the sample SD can be estimated — df reflects what's left over ($n-1$).
-   **F:** the ratio of two chi-square variables, each already carrying its own df — so the F distribution needs two df values, one for each piece of the ratio ($d_1, d_2$).

The pattern to notice across all three: **more degrees of freedom means more independent information, which means less uncertainty, which means the distribution's shape settles down and moves closer to Normal.** A chi-square with $k=2$ is sharply right-skewed; a chi-square with $k=50$ looks almost symmetric.
A $t$-distribution with $\nu=3$ has dramatically heavier tails than the Normal; by $\nu=100$ the two are nearly indistinguishable.
This is the same idea in every case, just counted differently depending on how the statistic was built.

*Why* degrees of freedom take the specific value they do and in particular, why estimating a mean from a sample "costs" exactly one degree of freedom, leaving $n-1$, is explained in detail in the chapter , once we've introduced how sample statistics are used to estimate unknown population parameters.
For now, it's enough to recognize df as a dial that controls how heavy-tailed or skewed these distributions are, and that larger df always means a distribution closer to something familiar and symmetric.
:::

## Chi-Square Distribution

### Definition

The distribution of the **sum of squares of** $k$ independent standard Normal variables; a special case of the Gamma distribution ($k$ degrees of freedom, rate $=1/2$).
Central to goodness-of-fit tests, tests of independence in contingency tables, and the sampling distribution of the sample variance.

$$X = \sum_{i=1}^{k} Z_i^2 \sim \chi^2_k$$

### Parameters, Mean, Variance

| Parameter | $k$ = degrees of freedom |
|-----------|--------------------------|
| Mean      | $k$                      |
| Variance  | $2k$                     |

### Properties

-   Right-skewed, especially for small $k$

-   Becomes more symmetric (approaching Normal) as $k$ increases

-   Always non-negative.

### Applications / Public Health Example

Chi-square test of independence for contingency tables (e.g., testing association between vaccination status and infection outcome); goodness-of-fit tests; constructing confidence intervals for a population variance.

### R Code and Graph

```{r chisq}
ggplot(data.frame(x = c(0, 20)), aes(x)) +
  stat_function(fun = dchisq, args = list(df = 4), geom = "area", fill = "chocolate", alpha = 0.5) +
  labs(title = "Chi-square(df = 4)", x = "X", y = "Density") + theme_minimal()
```

------------------------------------------------------------------------

## Student's t-Distribution

### Definition

Similar in shape to the standard Normal, but with **heavier tails**, used when estimating a population mean with an **unknown population variance** and a small sample size (the sample SD, $s$, is used in place of $\sigma$).

$$t = \frac{\bar{x}-\mu}{s/\sqrt{n}} \sim t_{n-1}$$

### Parameters, Mean, Variance

| Parameter | $\nu$ (degrees of freedom, typically $n-1$) |
|-----------|---------------------------------------------|
| Mean      | $0$ (for $\nu>1$)                           |
| Variance  | $\dfrac{\nu}{\nu-2}$ (for $\nu>2$)          |

### Properties

-   Symmetric, bell-shaped, centered at 0

-   Heavier tails than the standard Normal, especially for small $\nu$; converges to the standard Normal as $\nu\to\infty$ (in practice, $t$ and $Z$ are nearly identical for $\nu>30$).

### Applications / Public Health Example

The basis for the **one-sample and two-sample t-tests** and t-based confidence intervals for means (used throughout later chapters) when sample sizes are small and population SD is unknown, the standard situation in most clinical studies.

### R Code and Graph

```{r tdist}
ggplot(data.frame(x = c(-4, 4)), aes(x)) +
  stat_function(fun = dt, args = list(df = 5), aes(color = "t (df=5)")) +
  stat_function(fun = dnorm, aes(color = "Standard Normal")) +
  labs(title = "Student's t (df = 5) vs Standard Normal", x = "x", y = "Density", color = "") +
  theme_minimal()
```

------------------------------------------------------------------------

## F Distribution

### Definition

The distribution of the **ratio of two independent chi-square variables**, each divided by their own degrees of freedom, used to compare two variances, and as the basis for ANOVA (Analysis of Variance).

$$F = \frac{\chi^2_{d_1}/d_1}{\chi^2_{d_2}/d_2} \sim F_{d_1,d_2}$$

### Parameters, Mean, Variance

| Parameter | $d_1$ (numerator df), $d_2$ (denominator df)              |
|-----------|-----------------------------------------------------------|
| Mean      | $\dfrac{d_2}{d_2-2}$ (for $d_2>2$)                        |
| Variance  | Complex expression involving both $d_1,d_2$ (for $d_2>4$) |

### Properties

Right-skewed, always non-negative; shape depends on both degrees of freedom; approaches Normal-like shape as both $d_1,d_2$ grow large.

### Applications / Public Health Example

Comparing variability between two treatment groups (F-test for equal variances); the basis of **ANOVA**, used to compare means across three or more groups (e.g., comparing mean recovery time across three treatment protocols).

### R Code and Graph

```{r fdist}
ggplot(data.frame(x = c(0, 5)), aes(x)) +
  stat_function(fun = df, args = list(df1 = 5, df2 = 20), geom = "area", fill = "gold", alpha = 0.5) +
  labs(title = "F(df1 = 5, df2 = 20)", x = "F", y = "Density") + theme_minimal()
```

------------------------------------------------------------------------

## Comparison Table: Continuous Distributions

| Distribution | Range | Parameters | Mean | Variance | Shape | Typical Use |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| Uniform | $[a,b]$ | $a, b$ | $(a+b)/2$ | $(b-a)^2/12$ | Flat | Equally likely outcomes |
| Normal | $(-\infty,\infty)$ | $\mu,\sigma$ | $\mu$ | $\sigma^2$ | Symmetric, bell | Biological measurements, CLT |
| Standard Normal | $(-\infty,\infty)$ | none ($\mu=0,\sigma=1$) | 0 | 1 | Symmetric, bell | z-scores, standardization |
| Exponential | $[0,\infty)$ | $\lambda$ | $1/\lambda$ | $1/\lambda^2$ | Right-skewed | Waiting times, survival |
| Gamma | $[0,\infty)$ | $k,\lambda$ | $k/\lambda$ | $k/\lambda^2$ | Right-skewed | Time to $k$-th event, cost data |
| Chi-square | $[0,\infty)$ | $k$ (df) | $k$ | $2k$ | Right-skewed | Variance tests, goodness-of-fit |
| Student's t | $(-\infty,\infty)$ | $\nu$ (df) | 0 | $\nu/(\nu-2)$ | Symmetric, heavy tails | Small-sample mean inference |
| F | $[0,\infty)$ | $d_1,d_2$ (df) | $d_2/(d_2-2)$ | complex | Right-skewed | ANOVA, variance ratio tests |

## Decision Guide: Which Continuous Distribution Should I Use?

```{mermaid, echo=FALSE}
flowchart TD
    Q1{"What are you modeling?"} --> Q2["A single continuous<br/>measurement, symmetric,<br/>bell-shaped?"]
    Q1 --> Q3["Waiting time / time-to-event?"]
    Q1 --> Q4["A test statistic for<br/>inference?"]
    Q1 --> Q5["All outcomes<br/>equally likely in a range?"]
    Q2 -->|Yes| Normal["Use Normal<br/>(or Standard Normal if standardized)"]
    Q3 -->|"Time to 1st event"| Expo["Use Exponential"]
    Q3 -->|"Time to k-th event"| Gam["Use Gamma"]
    Q5 -->|Yes| Unif["Use Uniform"]
    Q4 --> Q6{"Comparing a mean,<br/>small n, sigma unknown?"}
    Q4 --> Q7{"Testing a variance or<br/>categorical association?"}
    Q4 --> Q8{"Comparing variances<br/>across 3+ groups (ANOVA)?"}
    Q6 -->|Yes| Tdist["Use Student's t"]
    Q7 -->|Yes| Chisq["Use Chi-square"]
    Q8 -->|Yes| Fdist["Use F distribution"]
```

## Summary

-   Continuous distributions use a **PDF** (density, area = probability) and **CDF** (cumulative probability) rather than a PMF.
-   The **Normal distribution** is central to biostatistics because of the Central Limit Theorem; the **Standard Normal** enables z-score based probability calculations.
-   The **Exponential** and **Gamma** distributions model waiting times, closely paralleling the discrete Geometric and Negative Binomial distributions.
-   The **Chi-square**, **Student's t**, and **F distributions** are not typically used to model raw data directly — they arise as **sampling distributions of test statistics**, and are the foundation of the hypothesis testing methods introduced in Sections 9-11 and developed throughout the rest of this book.
