---
title: "Estimation"
output:
  pdf_document:
    number_sections: yes
  html_document: default
editor_options:
  chunk_output_type: console
editor: 
  markdown: 
    wrap: sentence
---

```{r, include=FALSE}
knitr::opts_knit$set(root.dir = "D:/Epi Biostat with R/Biostat_with_R/")
data <- readxl::read_excel("D:/Epi Biostat with R/Biostat_with_R/Environment/Datasets/wcgs.xls")
```

## Estimation {#sec-estimation_chapter}

**Estimation** is the process of using sample data to infer the value of an unknown population parameter.
When estimating an unknown population parameter, different estimators may perform differently.
A **good estimator** is one that provides estimates that are close to the true population parameter and behaves reliably across repeated samples.
Four important properties commonly used to evaluate an estimator are **unbiasedness, efficiency, consistency, and sufficiency**.

### Unbiasedness

An estimator is said to be **unbiased** if its expected value is equal to the true population parameter.

If $\hat{\theta}$ is an estimator of a population parameter $\theta$, then $\hat{\theta}$ is unbiased if

$$
E(\hat{\theta}) = \theta.
$$

In other words, an unbiased estimator does not systematically overestimate or underestimate the population parameter.
Across many repeated samples, the average of the estimates will equal the true parameter.

For example, the sample mean $\bar{X}$ is an unbiased estimator of the population mean $\mu$:

$$
E(\bar{X}) = \mu.
$$

### 2. Efficiency

**Efficiency** refers to the precision of an estimator.
Among unbiased estimators of the same population parameter, the estimator with the **smaller variance** is generally considered more efficient.

For an estimator $\hat{\theta}$, its variance measures how much the estimates vary from sample to sample:

$$
\operatorname{Var}(\hat{\theta}).
$$

A smaller variance means that repeated estimates tend to be more closely clustered around the true population parameter.

For example, when estimating the population mean under the usual assumptions, the sample mean is more efficient than estimators such as the sample median because the sample mean generally has a smaller sampling variance.

> **In simple terms:** An efficient estimator provides more precise estimates with less sampling variability.

Efficiency should not be confused with unbiasedness.
An estimator can be unbiased but less efficient than another unbiased estimator.

### 3. Consistency

An estimator is **consistent** if it gets closer to the true population parameter as the sample size increases.

More formally, an estimator $\hat{\theta}_n$ is consistent for $\theta$ if

$$
\hat{\theta}_n \xrightarrow{P} \theta
\quad \text{as} \quad n \rightarrow \infty,
$$

where $\xrightarrow{P}$ denotes **convergence in probability**.

In practical terms, as more observations are collected, the probability that the estimator differs substantially from the true population parameter becomes increasingly small.

For example, the sample mean is a consistent estimator of the population mean.
As the sample size increases, the sample mean generally gets closer to the population mean.

> **Key idea:** Consistency concerns what happens as the sample size becomes very large.

### 4. Sufficiency

An estimator is **sufficient** if it contains all the information in the sample that is relevant to estimating a particular population parameter.

More formally, a statistic $T(X)$ is sufficient for a parameter $\theta$ if, once $T(X)$ is known, the remaining information in the sample provides no additional information about $\theta$.

The **factorization theorem** provides a useful mathematical criterion for determining whether a statistic is sufficient.
If the likelihood can be written as

$$
L(\theta \mid x) = g(T(x),\theta)h(x),
$$

then $T(x)$ is sufficient for $\theta$.

For example, for a random sample from a Poisson distribution with parameter $\lambda$, the sum

$$
T = \sum_{i=1}^{n}X_i
$$

is a sufficient statistic for $\lambda$.
Once the total number of observed events is known, the individual arrangement of those observations provides no additional information about $\lambda$.

> **Key idea:** Sufficiency means that the statistic captures all the information about the parameter contained in the sample.

### Summary of the Criteria

| Property | Main idea | Question it answers |
|------------------------|------------------------|------------------------|
| **Unbiasedness** | $E(\hat{\theta}) = \theta$ | Does the estimator systematically over- or underestimate? |
| **Efficiency** | Smaller variance among comparable estimators | How precise is the estimator? |
| **Consistency** | $\hat{\theta}$ approaches $\theta$ as $n$ increases | Does the estimator improve with larger samples? |
| **Sufficiency** | Retains all information about the parameter | Does the statistic contain all relevant information from the sample? |

::: callout-note
### Remember

**Unbiasedness** = correct on average\
**Efficiency** = precise\
**Consistency** = improves with increasing sample size\
**Sufficiency** = retains all relevant information
:::

## Point Estimation

A **point estimate** is a single numeric value, calculated from sample data, that serves as the "best guess" for a population parameter.

| Population Parameter                 | Point Estimator (Sample Statistic) |
|-------------------------------------|-----------------------------------|
| $\mu$ (population mean)              | $\bar{x}$ (sample mean)            |
| $\sigma^2$ (population variance)     | $s^2$ (sample variance)            |
| $\pi$ or $P$ (population proportion) | $\hat{p}$ (sample proportion)      |

**Desirable properties of a good point estimator:** unbiasedness (its expected value equals the true parameter, e.g., $E(\bar{x})=\mu$), consistency (it converges to the true parameter as $n\to\infty$), and efficiency (it has the smallest possible variance among unbiased estimators).

::: callout-important
A point estimate alone gives **no information about its precision or reliability**.
A sample mean of $\bar{x}=126$ mmHg conveys nothing about how much uncertainty surrounds that estimate.
Was it based on 10 patients or 10,000?
**Interval estimation** solves this problem.
:::

## Standard Error

The **standard error (SE)** is the standard deviation of the sampling distribution of a statistic — it quantifies how much a sample statistic (like $\bar{x}$) would vary from sample to sample if we repeated the sampling process many times.

$$SE(\bar{x}) = \frac{s}{\sqrt{n}} \qquad SE(\hat{p}) = \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

::: callout-tip
Do not confuse **standard deviation (SD)** with **standard error (SE)**.
The SD describes variability of **individual observations** in the data; the SE describes variability of the **sample statistic itself** (e.g., the mean) across repeated samples.
As $n$ increases, SD stays roughly constant (it reflects the population), but SE **shrinks** (larger samples give more precise/stable estimates of the mean), because $SE = s/\sqrt{n}$.
:::

## Interval Estimation and the Confidence Interval {#sec-confidence_interval}

### Definition

A **confidence interval (CI)** is a range of values, calculated from sample data, that is likely to contain the true population parameter, along with a stated **confidence level** (e.g., 95%) reflecting the long-run reliability of the method used to construct it.

### General Formula

$$\text{Point estimate} \pm (\text{Critical value}) \times (\text{Standard error})$$ where,

$Margin.of.error(ME)=critical.value \times standard.error$

### Confidence Interval for a Mean

When $\sigma$ is unknown, we use $t$ statistics,

$$\bar{x} \pm t_{\alpha/2, n-1} \times \frac{s}{\sqrt{n}}$$

### Confidence Interval for a Proportion

$$\hat{p} \pm z_{\alpha/2}\times\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

## Margin of Error

The **margin of error (ME)** is the "plus-or-minus" part of the confidence interval: $$ME = (\text{Critical value})\times(\text{Standard error})$$

A larger margin of error means a wider, less precise interval; margin of error decreases as sample size increases (since $SE$ shrinks) and increases as the desired confidence level increases (a higher critical value is needed for more confidence).

## Confidence Level and Its Interpretation

Common confidence levels and their corresponding critical $z$-values (for large samples/known $\sigma$):

| Confidence Level | $\alpha$ | Critical $z_{\alpha/2}$ |
|------------------|----------|-------------------------|
| 90%              | 0.10     | 1.645                   |
| 95%              | 0.05     | 1.96                    |
| 99%              | 0.01     | 2.576                   |

::: callout-important
**Correct interpretation of a 95% CI:** "If we were to repeat this study many times, drawing a new random sample each time and constructing a CI the same way, approximately 95% of those intervals would contain the true population parameter." **It is incorrect** to say "there is a 95% probability that the true parameter lies in this particular interval".
Under the frequentist framework, the true parameter is a fixed (though unknown) constant, and any single computed interval either contains it or does not (this is closer to the Bayesian *credible interval* interpretation).
:::

## Worked Example (Confidence interval for a mean)

Using our earlier sample: $n=36$ patients, $\bar{x}=126$ mmHg, $s=12$ mmHg.
Construct a 95% CI for the true mean blood pressure.

**Step 1:** Compute the standard error: $SE = s/\sqrt{n} = 12/\sqrt{36} = 2.0$.

**Step 2:** Find the critical value: for a 95% CI with $n-1=35$ df

$t_{0.025,35}\approx 2.03$ (close to the $z$ value of 1.96 for large $n$).

**Step 3:** Compute the margin of error: $ME = 2.03\times2.0=4.06$.

**Step 4:** Construct the interval: $$126 \pm 4.06 = (121.9,\ 130.1) \text{ mmHg}$$

**Interpretation:** We are 95% confident that the true mean blood pressure of the population from which this sample was drawn lies between 121.9 and 130.1 mmHg.
Note this interval **includes** 130 mmHg, the value tested in Section 9's hypothesis test — consistent with that test's borderline result (p = 0.046, just below 0.05, so technically still significant, illustrating that CIs and hypothesis tests can appear to be right at the boundary of agreement/disagreement when a result is only marginally significant).

## Determine the confidence interval around mean in R

We can use `meanCI()` from **DescTools** to calculate CI around mean in R.

```{r}
# CI around mean for problem above
mean_sbp <- 126
sd_sbp <- 12
n <- 36

se <- sd_sbp / sqrt(n)
t_crit <- qt(0.975, df = n - 1)

lower <- mean_sbp - t_crit * se
upper <- mean_sbp + t_crit * se

print(paste("Confidence interval around", mean_sbp, "is", c(lower, upper)))

# CI around mean for dataset
data <- readxl::read_excel("Section2/Categorical data analysis/Datasets/wcgs.xls")

#CI around mean  for data
DescTools::MeanCI(x = data$bmi, sd = sd(data$bmi), conf.level = 0.95,sides = "two.sided") 
```

## Worked Example: Confidence Interval for Proportion

In a survey of $n=400$ adults, 92 report having been vaccinated against influenza this season.
Construct a 95% CI for the true population proportion vaccinated.

**Step 1:** $\hat{p} = 92/400 = 0.23$.

**Step 2:** $SE(\hat{p}) = \sqrt{\dfrac{0.23(1-0.23)}{400}} = \sqrt{\dfrac{0.1771}{400}} = \sqrt{0.000443}=0.02104$

**Step 3:** $ME = 1.96\times0.02104=0.0412$

**Step 4:** $$0.23 \pm 0.0412 = (0.189,\ 0.271) \text{ or } (18.9\%,\ 27.1\%)$$

**Interpretation:** We are 95% confident that the true influenza vaccination rate in the population is between 18.9% and 27.1%.

## Determine the confidence interval around proportion in R

Confidence intervals can be produced for either binomial or multinomial proportions.

The *`binom.test()`* function in the native *stats* package will provide the Clopper-Pearson confidence interval for a binomial proportion.
Other methods for a binomial proportion are provided by the *`BinomCI()`* function in the *DescTools* package, as well as by various functions in the *PropCIs* package.
Different methods available for calculating CI around binary variable include : `"wilson", "wald", "agresti-coull", "jeffreys", "modified wilson", "modified jeffreys","clopper-pearson", "arcsine", "logit", "witting", "pratt"`

Confidence intervals for multinomial proportions can be produced with the *`MultinomCI()`* function in the *DescTools* package.

```{r}
# CI for proportion calculation manually for problem above
p <- 60 / 200
k <- p + c(-1, 1) * qnorm(0.975) * sqrt(p * (1 - p) / 200)

print(paste("The confidence interval around proportion", p, "is", round(k[1],2), "and ", round(k[2],2) ))


# # CI for proportion calculation for dataset
data <- readxl::read_excel("Section2/Categorical data analysis/Datasets/wcgs.xls")

# Confidence interval for binary variable  
a<- addmargins(table(data$chd69))  
# In the above table No is 2897 and yes is 257 and total of 3154 

DescTools::BinomCI(x = a[2],n = a[3], method = "clopper")*100 # we can select different methods depending on need  

# Confidence interval around multinomial variable 
b<- addmargins(table(data$agec))  
# In the above table for each categories 35-40, 41-45, 46-50, 51-55 and 56-60, number of observations are 543, 1091, 750, 528, and 242 respectively

DescTools::MultinomCI( x = b, method = "goodman")*100 # we can select different methods depending on need  
```

## Confidence Interval vs. Prediction Interval

| Feature | Confidence Interval | Prediction Interval |
|------------------------|------------------------|------------------------|
| Estimates | A **population parameter** (e.g., the true mean $\mu$) | A **single future/individual observation** |
| Width | Narrower | **Always wider** than the CI for the same data and confidence level |
| Accounts for | Uncertainty in estimating the parameter only | Uncertainty in estimating the parameter **plus** the natural variability of individual observations |
| Formula (mean) | $\bar{x}\pm t_{\alpha/2,n-1}\dfrac{s}{\sqrt n}$ | $\bar{x}\pm t_{\alpha/2,n-1}\, s\sqrt{1+\dfrac{1}{n}}$ |
| Example question | "What is the plausible range for the *average* blood pressure of all patients like this?" | "What is the plausible range of blood pressure for *one new* patient?" |

::: callout-tip
A useful way to remember the distinction: a **confidence interval** narrows toward zero width as $n\to\infty$ (we can estimate the population mean with near-perfect precision given enough data), but a **prediction interval** never shrinks below the natural variability of individual observations ($s$), no matter how large $n$ becomes, individual patients will always vary, even if we know the population mean exactly.
:::

## Summary

-   **Point estimation** provides a single best-guess value for a parameter; **interval estimation** adds a range reflecting the estimate's precision.
-   The **standard error** measures the variability of the sample statistic itself (distinct from the SD, which measures variability of individual observations) and shrinks as $n$ increases.
-   A **confidence interval** = point estimate ± (critical value × standard error); wider intervals reflect either smaller sample sizes or higher confidence levels.
-   The correct interpretation of a 95% CI concerns the **long-run performance of the method**, not the probability that a specific interval contains the true parameter.
-   A **prediction interval** is always wider than a confidence interval because it must account for individual-level variability in addition to parameter uncertainty.

## References
