---
title: "Association Between Two Numeric Variables"
output:
  pdf_document:
    number_sections: yes
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_knit$set(root.dir = "D:/Epi Biostat with R/Biostat_with_R/")
```

```{r, include=FALSE, warning=FALSE,message=FALSE, echo=FALSE}
#load libraries and datasets
library(tidyverse)
library(car)        # leveneTest(), Anova() (Type II/III SS)
library(DescTools)  # CI helpers used later in the chapter
library(kableExtra)
library(corrplot)
library(knitr)
library(psych)

wcgs <- readxl::read_excel("Section2/Categorical data analysis/Datasets/wcgs.xls")
```

# Learning Objectives {.unnumbered}

By the end of this chapter, you should be able to:

1.  Explain what it means for two numeric (continuous) variables to be "associated," and describe that association using a scatter diagram.
2.  State the definition, assumptions, properties, and correct interpretation of the **Pearson product-moment correlation coefficient** ($r$).
3.  Calculate $r$ **by hand** from raw data, test $H_0: \rho = 0$, and construct a confidence interval for $\rho$.
4.  State the definition, assumptions, properties, and correct interpretation of the **Spearman rank correlation coefficient** ($r_s$ or $\rho_s$).
5.  Calculate $r_s$ **by hand**, including the correction needed when tied ranks are present.
6.  Reproduce both coefficients in **R**, using the Western Collaborative Group Study (**WCGS**) data set as a worked example, and correctly choose between Pearson and Spearman for a given pair of variables.
7.  Recognise common pitfalls in interpreting correlation coefficients.

------------------------------------------------------------------------

# Association Between Two Numeric Variables

In epidemiology and clinical research we frequently ask whether two **quantitative (numeric)** characteristics measured on the same individuals tend to move together — for example, does systolic blood pressure rise with age? Does body weight track with height? Does serum cholesterol relate to blood pressure?

**Association** (or *correlation*, used loosely) between two numeric variables describes the tendency of one variable to change systematically as the other changes. It has three components that should always be examined together:

-   **Direction** — do the variables move in the same direction (*positive* association) or opposite directions (*negative* association)?
-   **Strength** — how closely do the points cluster around a discernible trend?
-   **Form** — is the trend linear, curvilinear (monotonic but not straight), or is there no discernible pattern?

## The scatter diagram

The scatter diagram (scatterplot) is the essential first step in *any* correlation analysis — a numeric summary should never be trusted without first looking at the data.

```{r}
#| label: fig-scatter-intro
#| fig-cap: "Height and weight in the WCGS cohort (each point is one man)."

ggplot(wcgs, aes(x = height, y = weight)) +
  geom_point(alpha = 0.25, colour = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick") +
  labs(x = "Height (inches)", y = "Weight (lb)",
       title = "Height vs. weight, WCGS cohort (n = 3,154)") +
  theme_minimal(base_size = 12)
```

Two correlation coefficients dominate practice, and this chapter covers both:

|   | **Pearson** $r$ | **Spearman** $r_s$ |
|----|----|----|
| Measures | Strength of **linear** association | Strength of **monotonic** association |
| Uses | Raw values | **Ranks** of the values |
| Best suited to | Approximately normal, linearly related continuous data | Ordinal data, non-normal data, data with outliers, or monotonic-but-curved relationships |

------------------------------------------------------------------------

## Pearson Product-Moment Correlation Coefficient

### Definition {#sec-ppmc}

For a population, the Pearson correlation coefficient is

$$
\rho = \frac{\operatorname{Cov}(X,Y)}{\sigma_X\,\sigma_Y}
$$

For a sample of $n$ paired observations $(x_i, y_i)$, it is estimated by

$$
r = \frac{\displaystyle\sum_{i=1}^{n}(x_i-\bar x)(y_i-\bar y)}
        {\sqrt{\displaystyle\sum_{i=1}^{n}(x_i-\bar x)^2 \; \sum_{i=1}^{n}(y_i-\bar y)^2}}
$$ {#eq-pearson-deviation}

an equivalent **computational (raw-score) formula**, more convenient for hand calculation, is

$$
r = \frac{n\sum x_iy_i - \left(\sum x_i\right)\left(\sum y_i\right)}
        {\sqrt{\left[n\sum x_i^2-\left(\sum x_i\right)^2\right]\left[n\sum y_i^2-\left(\sum y_i\right)^2\right]}}
$$ {#eq-pearson-raw}

## Assumptions

1.  Both $X$ and $Y$ are **continuous (interval/ratio) numeric** variables.
2.  The relationship between $X$ and $Y$ is (approximately) **linear**.
3.  For inference (hypothesis tests / CIs), the pairs $(X,Y)$ come from a **bivariate normal distribution** — in practice this is usually checked by inspecting the marginal distributions of $X$ and $Y$ for gross non-normality and skew.
4.  Observations are **independent** of one another.
5.  No excessively influential **outliers**.
6.  Variance of $Y$ is reasonably constant across the range of $X$ (homoscedasticity).

## Properties

-   $-1 \le r \le 1$; $r=1$ is perfect positive linear association, $r=-1$ perfect negative linear association, $r=0$ no *linear* association.
-   $r$ is **unit-free** (dimensionless) and **symmetric**: $r_{xy} = r_{yx}$.
-   $r$ measures only **linear** association — a strong curved relationship can produce a small $r$.
-   $r$ is sensitive to **outliers** and to **restriction of range**.
-   $r^2$ (the **coefficient of determination**) is the proportion of variance in $Y$ "explained" by the linear relationship with $X$.
-   Correlation does **not** imply causation.

## A rule-of-thumb interpretation scale

| $|r|$       | Strength of linear association |
|-------------|--------------------------------|
| 0.00 – 0.09 | Negligible                     |
| 0.10 – 0.39 | Weak                           |
| 0.40 – 0.69 | Moderate                       |
| 0.70 – 0.89 | Strong                         |
| 0.90 – 1.00 | Very strong                    |

These cut-points (adapted from conventions such as Cohen, 1988) are a rough guide only; the meaningfulness of a given $r$ always depends on the substantive context.

## Hypothesis test for $\rho$

To test $H_0: \rho = 0$ versus $H_1: \rho \ne 0$:

$$
t = r\sqrt{\frac{n-2}{1-r^2}} \qquad \sim \; t_{(n-2)} \text{ under } H_0
$$ {#eq-pearson-t}

## Confidence interval for $\rho$ (Fisher's $z$-transformation)

Because the sampling distribution of $r$ is skewed (except when $\rho=0$), Fisher's $z$-transformation is used:

$$
z' = \tfrac12 \ln\!\left(\frac{1+r}{1-r}\right), \qquad SE(z') = \frac{1}{\sqrt{n-3}}
$$

A $95\%$ CI for $z'$ is $z' \pm 1.96 \times SE(z')$; each limit is back-transformed with

$$
r = \frac{e^{2z'}-1}{e^{2z'}+1}
$$

------------------------------------------------------------------------

# Manual (Hand) Calculation of Pearson $r$

We illustrate the arithmetic with a small, reproducible random sample of $n=10$ men drawn from the WCGS data, using **height** (inches) and **weight** (lb).

```{r}
#| label: pearson-manual-sample
set.seed(123)
idx <- sample(seq_len(nrow(wcgs)), 10)
sub <- as.data.frame(wcgs[idx, c("id", "height", "weight")])
names(sub) <- c("id", "x_height", "y_weight")
kable(sub, caption = "Random subsample (n = 10) used for the manual Pearson calculation")
```

## Step 1: Compute the deviation table

For each subject we need $x-\bar x$, $y-\bar y$, their product, and their squares.

```{r}
#| label: pearson-manual-table
x <- sub$x_height; y <- sub$y_weight
n <- length(x)
xbar <- mean(x); ybar <- mean(y)

tab <- data.frame(
  id     = sub$id,
  x      = x,
  y      = y,
  dx     = round(x - xbar, 2),
  dy     = round(y - ybar, 2),
  dx_dy  = round((x - xbar) * (y - ybar), 2),
  dx2    = round((x - xbar)^2, 2),
  dy2    = round((y - ybar)^2, 2)
)
kable(tab, col.names = c("ID","x (height)","y (weight)","$x-\\bar x$","$y-\\bar y$",
                          "$(x-\\bar x)(y-\\bar y)$","$(x-\\bar x)^2$","$(y-\\bar y)^2$"))

cat("Column sums:\n",
    " sum(x) =", sum(x), "  sum(y) =", sum(y), "\n",
    " sum[(x-xbar)(y-ybar)] =", sum(tab$dx_dy), "\n",
    " sum[(x-xbar)^2]       =", sum(tab$dx2), "\n",
    " sum[(y-ybar)^2]       =", sum(tab$dy2), "\n")
```

## Step 2: Apply the deviation formula

$$
\bar x = \frac{694}{10}=69.4 \qquad \bar y=\frac{1750}{10}=175.0
$$

$$
r=\frac{\sum(x-\bar x)(y-\bar y)}{\sqrt{\sum(x-\bar x)^2 \sum(y-\bar y)^2}}
 =\frac{831.0}{\sqrt{72.4 \times 16624}}
 =\frac{831.0}{\sqrt{1{,}203{,}977.6}}
 =\frac{831.0}{1097.26}
$$

```{r}
#| label: pearson-manual-r
r_manual <- sum(tab$dx_dy) / sqrt(sum(tab$dx2) * sum(tab$dy2))
r_manual
```

**By hand,** $r = 0.757$ — a strong, positive linear association between height and weight in this subsample.

### Cross-check with the computational (raw-score) formula

```{r}
#| label: pearson-manual-raw
Sx <- sum(x); Sy <- sum(y); Sxy <- sum(x*y); Sx2 <- sum(x^2); Sy2 <- sum(y^2)
r_raw <- (n*Sxy - Sx*Sy) / sqrt((n*Sx2 - Sx^2) * (n*Sy2 - Sy^2))
data.frame(Sx, Sy, Sxy, Sx2, Sy2, r_raw)
```

Both formulas agree exactly, as they must algebraically.

## Step 3: Test $H_0: \rho = 0$

$$
t = r\sqrt{\frac{n-2}{1-r^2}} = 0.757\sqrt{\frac{8}{1-0.757^2}}
$$

```{r}
#| label: pearson-manual-ttest
t_manual  <- r_manual * sqrt((n - 2) / (1 - r_manual^2))
df_manual <- n - 2
p_manual  <- 2 * pt(-abs(t_manual), df = df_manual)
data.frame(t = t_manual, df = df_manual, p_value = p_manual)
```

With $t = 3.28$ on $8$ df, $p = 0.011$: we reject $H_0$ at the $5\%$ level and conclude there is a statistically significant positive linear association between height and weight in this small sample.

## Step 4: Verify against R's built-in function

```{r}
#| label: pearson-manual-verify
cor.test(x, y, method = "pearson")
```

The hand-calculated $r$, $t$, $df$, and $p$-value match R's `cor.test()` output exactly — confirming the manual arithmetic.

------------------------------------------------------------------------

# Spearman Rank Correlation Coefficient

## When to use Spearman instead of Pearson {#sec-spearman}

Spearman's coefficient should be preferred over Pearson's when:

-   The variables are **ordinal** rather than truly numeric (e.g., pain scores, tumour grade).
-   The data are numeric but **markedly non-normal or skewed** (Pearson's inferential validity is weakened).
-   There are **influential outliers** that would distort $r$.
-   The relationship is **monotonic but not linear** (consistently increasing or decreasing, but curved) — Spearman still detects this fully, whereas Pearson under-estimates it.

## Definition

Spearman's rank correlation is simply the Pearson correlation coefficient computed on the **ranks** of $X$ and $Y$ rather than on their raw values. If there are no tied ranks, this simplifies to the well-known shortcut formula:

$$
r_s = 1-\frac{6\sum_{i=1}^n d_i^2}{n(n^2-1)}, \qquad d_i = \operatorname{rank}(x_i)-\operatorname{rank}(y_i)
$$ {#eq-spearman}

**Tied values** are given the **average of the ranks** they would otherwise occupy. When ties are present, @eq-spearman is only an *approximation*; the exact value is obtained by computing the ordinary Pearson formula (@eq-pearson-deviation) on the rank-transformed data. Software (including R) uses this ties-corrected version by default.

## Assumptions and properties

-   Requires only that the data be at least **ordinal**, and that the relationship be **monotonic**.
-   $-1 \le r_s \le 1$, same interpretation of sign and magnitude as Pearson's $r$, but describing a **monotonic** rather than strictly linear trend.
-   Far more **robust to outliers** than Pearson's $r$, because extreme values are pulled in by ranking.
-   Hypothesis test of $H_0: \rho_s = 0$ can use the same $t$-approximation as Pearson's (@eq-pearson-t), substituting $r_s$ for $r$, for moderate/large $n$; exact permutation-based $p$-values are used for small $n$ without ties.

------------------------------------------------------------------------

# Manual (Hand) Calculation of Spearman $r_s$

We again draw a reproducible random subsample of $n=10$ men (this time keeping only those with a recorded cholesterol value) and examine **total cholesterol** (`chol`, mg/dL) against **systolic blood pressure** (`sbp`, mmHg) — two variables that are typically right-skewed, making Spearman a more defensible choice than Pearson.

```{r}
#| label: spearman-manual-sample
set.seed(321)
idx2 <- sample(which(!is.na(wcgs$chol)), 10)
sub2 <- as.data.frame(wcgs[idx2, c("id", "chol", "sbp")])
kable(sub2, caption = "Random subsample (n = 10, complete cholesterol) used for the manual Spearman calculation")
```

## Step 1: Rank each variable separately

Ranks are assigned from lowest (rank 1) to highest; **tied values receive the average of the ranks they span**. Notice that `sbp` = 120 occurs three times (ranks 3, 4, 5 → averaged to 4).

```{r}
#| label: spearman-manual-ranks
cx <- sub2$chol; cy <- sub2$sbp
rx <- rank(cx); ry <- rank(cy)
d  <- rx - ry
d2 <- d^2

tab2 <- data.frame(id = sub2$id, chol = cx, sbp = cy,
                    rank_chol = rx, rank_sbp = ry, d = d, d2 = d2)
kable(tab2, col.names = c("ID","Cholesterol","SBP","Rank (chol)","Rank (sbp)","$d$","$d^2$"))
cat("Sum of d^2 =", sum(d2), "\n")
```

## Step 2: Apply the shortcut formula (ignoring ties)

$$
r_s \approx 1-\frac{6\sum d_i^2}{n(n^2-1)} = 1-\frac{6(114.5)}{10(100-1)} = 1-\frac{687}{990}
$$

```{r}
#| label: spearman-manual-shortcut
n2 <- length(cx)
rho_shortcut <- 1 - (6 * sum(d2)) / (n2 * (n2^2 - 1))
rho_shortcut
```

This gives $r_s \approx 0.306$. **However**, because `sbp` contains tied values, this shortcut formula is only approximate.

## Step 3: Exact value: Pearson's formula applied to the ranks

```{r}
#| label: spearman-manual-exact
rho_exact <- cor(rx, ry, method = "pearson")
rho_exact
```

The tie-corrected (exact) value is $r_s = 0.295$ — close to, but not identical to, the uncorrected shortcut (0.306). **Rule of thumb: with any tied values, always use the Pearson-on-ranks method (or software) rather than the raw** $d^2$ shortcut.

## Step 4: Verify against R's built-in function

```{r}
#| label: spearman-manual-verify
cor.test(cx, cy, method = "spearman")
```

R's `cor.test()` reports $r_s = 0.295$, matching our tie-corrected hand calculation exactly (R issues a warning that an *exact* $p$-value cannot be computed because of ties, and instead reports an asymptotic $t$-approximation $p$-value). With $n=10$ and $r_s=0.295$, $p = 0.41$ — no statistically significant monotonic association is detected in this small subsample.

------------------------------------------------------------------------

# Demonstration in R Using the Full WCGS Data Set

## About the data

The **Western Collaborative Group Study (WCGS)** was a prospective cohort study of coronary heart disease (CHD) among middle-aged American men. The extract used here (`wcgs.xls`) contains **`r nrow(wcgs)`** subjects and the following key numeric variables:

```{r}
#| label: tbl-dictionary
#| tbl-cap: "Selected WCGS variables used in this chapter"
dict <- data.frame(
  Variable = c("age","height","weight","bmi","sbp","dbp","chol","ncigs","chd69","behpat"),
  Description = c(
    "Age (years)",
    "Height (inches)",
    "Weight (pounds)",
    "Body mass index (kg/m$^2$)",
    "Systolic blood pressure (mmHg)",
    "Diastolic blood pressure (mmHg)",
    "Total serum cholesterol (mg/dL)",
    "Cigarettes smoked per day",
    "Coronary heart disease event by end of follow-up (Yes/No)",
    "Behaviour pattern (A1, A2, B3, B4)"
  )
)
kable(dict)
```

## Step 1: Import the data

```{r}
#| label: import
#| eval: false
library(readxl)
wcgs <- read_excel("wcgs.xls")
str(wcgs)
```

```{r}
#| label: import-summary
summary(wcgs[, c("age","height","weight","bmi","sbp","dbp","chol","ncigs")])
```

## Step 2: Visualise the relationships first

```{r}
#| label: fig-scatter-two
#| fig-cap: "Two candidate relationships: (left) height vs. weight — roughly linear; (right) cholesterol vs. systolic BP — noisier, right-skewed variables."
#| fig-subcap:
#|   - "Height vs. weight"
#|   - "Cholesterol vs. SBP"
#| layout-ncol: 2
ggplot(wcgs, aes(height, weight)) +
  geom_point(alpha = 0.25, colour = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick") +
  labs(x = "Height (in)", y = "Weight (lb)") +
  theme_minimal()

ggplot(wcgs, aes(chol, sbp)) +
  geom_point(alpha = 0.25, colour = "darkorange3") +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick") +
  labs(x = "Cholesterol (mg/dL)", y = "SBP (mmHg)") +
  theme_minimal()
```

## Step 3: Check the normality assumption (needed for Pearson inference)

```{r}
#| label: fig-qq
#| fig-cap: "Normal Q-Q plots: height and weight are close to normal; cholesterol and SBP are right-skewed."
#| fig-subcap:
#|   - "Height"
#|   - "Weight"
#|   - "Cholesterol"
#|   - "SBP"
#| layout-ncol: 2
qqnorm(wcgs$height, main = "Height"); qqline(wcgs$height, col = "red")
qqnorm(wcgs$weight, main = "Weight"); qqline(wcgs$weight, col = "red")
qqnorm(wcgs$chol,   main = "Cholesterol"); qqline(wcgs$chol, col = "red")
qqnorm(wcgs$sbp,    main = "SBP"); qqline(wcgs$sbp, col = "red")
```

```{r}
#| label: shapiro
shapiro.test(wcgs$height)
shapiro.test(wcgs$chol)
```

With $n = 3{,}154$, the Shapiro-Wilk test is extremely sensitive to trivial departures from normality (essentially all large samples reject $H_0$); the Q-Q plots are the more useful diagnostic here. Height is close to normal; cholesterol and SBP show the right skew typical of biological "positive-only" measurements, which is a caution for Pearson inference — and part of why we illustrate Spearman with this pair.

## Step 4: Pearson correlation

```{r}
#| label: pearson-full-hw
cor.test(wcgs$height, wcgs$weight, method = "pearson")
```

**Interpretation:** $r = 0.533$ ($95\%$ CI: $0.507$ to $0.557$), $t(3152) = 35.4$, $p < 0.001$. There is a **moderate, highly statistically significant, positive linear association** between height and weight: taller men tend to weigh more. $r^2 = 0.284$, i.e., height linearly "explains" about $28\%$ of the variability in weight.

```{r}
#| label: pearson-full-cholsbp
cor.test(wcgs$chol, wcgs$sbp, method = "pearson")
```

Pearson's $r = 0.123$ for cholesterol vs. SBP — a weak association — but recall both variables are right-skewed, so this Pearson estimate may be distorted by a handful of high values.

## Step 5: Spearman correlation

```{r}
#| label: spearman-full-cholsbp
cor.test(wcgs$chol, wcgs$sbp, method = "spearman", exact = FALSE)
```

**Interpretation:** $r_s = 0.138$, $p < 0.001$. Because $n$ is large, even this weak monotonic association is statistically significant — a reminder that with large samples, statistical significance and practical/clinical importance are not the same thing. Note `exact = FALSE` is required here because `chol` contains tied values and is large, so R falls back on the asymptotic $t$-approximation for the $p$-value.

```{r}
#| label: spearman-full-hw
cor.test(wcgs$height, wcgs$weight, method = "spearman", exact = FALSE)
```

For height and weight, Pearson ($r=0.533$) and Spearman ($r_s=0.529$) agree closely — expected, since this relationship is close to linear and both variables are reasonably symmetric.

## Step 6: A correlation matrix and correlation plot

It is often useful to examine several numeric variables at once.

```{r}
#| label: fig-corrmat
#| fig-cap: "Pearson correlation matrix of key WCGS numeric variables."
vars <- c("age","height","weight","bmi","sbp","dbp","chol","ncigs")
M <- cor(wcgs[, vars], use = "pairwise.complete.obs", method = "pearson")
round(M, 2)

corrplot(M, method = "circle", type = "upper", addCoef.col = "black",
         number.cex = 0.7, tl.col = "black", tl.srt = 45,
         title = "Pearson correlations, WCGS numeric variables",
         mar = c(0,0,2,0))
```

```{r}
#| label: fig-pairspanel
#| fig-cap: "Pairwise scatterplots, distributions, and Pearson correlations for a subset of WCGS variables."
psych::pairs.panels(wcgs[, c("age","height","weight","sbp","chol")],
                     method = "pearson", hist.col = "lightblue",
                     density = TRUE, ellipses = FALSE)
```

## Step 7: Summary table, Pearson and Spearman side by side

```{r}
#| label: tbl-summary-both
pairs_list <- list(c("height","weight"), c("age","sbp"), c("chol","sbp"), c("bmi","sbp"))

results <- do.call(rbind, lapply(pairs_list, function(p) {
  x <- wcgs[[p[1]]]; y <- wcgs[[p[2]]]
  ok <- complete.cases(x, y)
  pear <- cor.test(x[ok], y[ok], method = "pearson")
  spea <- cor.test(x[ok], y[ok], method = "spearman", exact = FALSE)
  data.frame(Variable_1 = p[1], Variable_2 = p[2], n = sum(ok),
             Pearson_r = round(unname(pear$estimate), 3),
             Pearson_p = signif(pear$p.value, 3),
             Spearman_rs = round(unname(spea$estimate), 3),
             Spearman_p = signif(spea$p.value, 3))
}))
kable(results)
```

------------------------------------------------------------------------

# Pearson vs. Spearman: A Side-by-Side Comparison

| Feature | Pearson $r$ | Spearman $r_s$ |
|----|----|----|
| Data type required | Numeric (interval/ratio) | Ordinal or numeric |
| Detects | Linear association | Any monotonic association |
| Uses | Actual values | Ranks of the values |
| Sensitive to outliers? | Yes, strongly | No, largely robust |
| Requires (approx.) normality for inference? | Yes | No |
| Computation | Covariance / product of SDs | Pearson formula applied to ranks |
| Typical use case | Height vs. weight, dose vs. response (linear) | Skewed labs (e.g., triglycerides), ordinal scores, monotone-but-curved relationships |

------------------------------------------------------------------------

# Common Pitfalls in Interpreting Correlation

-   **Correlation is not causation.** A significant $r$ or $r_s$ establishes association, not that one variable causes changes in the other; confounding, reverse causation, and chance are all still possible explanations.
-   **Non-linearity.** Pearson's $r$ can be close to zero even when $X$ and $Y$ are strongly related, if the relationship is curved (e.g., U-shaped) — always plot the data first.
-   **Outliers and leverage points.** A single extreme observation can inflate or deflate Pearson's $r$ substantially; Spearman is more robust in this situation.
-   **Restriction of range.** Correlation estimated within a narrow range of $X$ (e.g., only healthy adults) will understate the true association in the full population.
-   **Ecological correlation.** Correlating group-level averages (e.g., country-level means) can produce misleadingly strong correlations that do not hold at the individual level (ecological fallacy).
-   **Statistical vs. clinical significance.** With very large $n$ (as in WCGS, $n>3{,}000$), even trivially small correlations become "statistically significant." Always report and interpret the magnitude of $r$/$r_s$, not just the $p$-value.
-   **Multiple testing.** Scanning many pairwise correlations (e.g., a full correlation matrix) inflates the chance of false positives; interpret exploratory matrices cautiously and confirm key findings with pre-specified hypotheses.

------------------------------------------------------------------------

# Summary

-   The **scatter diagram** is the starting point for any correlation analysis.
-   **Pearson's** $r$ quantifies the strength and direction of a **linear** relationship between two numeric variables; it assumes approximate bivariate normality and is sensitive to outliers.
-   **Spearman's** $r_s$ quantifies the strength and direction of a **monotonic** relationship using ranks; it requires only ordinal data, handles skew and outliers gracefully, and reduces to a simple $d^2$ shortcut formula when there are no tied ranks (with a small correction needed when ties are present).
-   Both coefficients range from $-1$ to $+1$, are tested against $H_0: \rho = 0$ using a $t$-based statistic, and should always be reported together with sample size, a confidence interval where possible, and — most importantly — a plot of the raw data.
-   In the WCGS data, height and weight show a moderate, near-linear positive association ($r \approx r_s \approx 0.53$), while cholesterol and systolic blood pressure show only a weak monotonic association ($r=0.12$, $r_s=0.14$) despite reaching statistical significance because of the large sample size.

------------------------------------------------------------------------

# References {.unnumbered}

-   Rosner B. *Fundamentals of Biostatistics*. 8th ed. Cengage Learning.
-   Daniel WW, Cross CL. *Biostatistics: A Foundation for Analysis in the Health Sciences*. 11th ed. Wiley.
-   Bland M. *An Introduction to Medical Statistics*. 4th ed. Oxford University Press.
-   Woodward M. *Epidemiology: Study Design and Data Analysis*. 3rd ed. Chapman & Hall/CRC.
-   Cohen J. *Statistical Power Analysis for the Behavioral Sciences*. 2nd ed. Lawrence Erlbaum Associates, 1988.
-   Rosenman RH, et al. "A predictive study of coronary heart disease: the Western Collaborative Group Study." *JAMA*. 1964;189:15–22. (Original WCGS study.)
