---
title: "Comparing a Sample Mean to a Fixed Value"
output:
  pdf_document:
    number_sections: yes
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_knit$set(root.dir = "D:/Epi Biostat with R/Biostat_with_R/")
```

```{r}
#| label: setup
#| include: false
library(readxl)
library(ggplot2)
library(knitr)
library(car)
 
options(digits = 4)
wcgs <- readxl::read_excel("Section2/Categorical data analysis/Datasets/wcgs.xls")
wcgs$behpat <- factor(wcgs$behpat, levels = c("A1", "A2", "B3", "B4"))
wcgs$dibpat <- factor(wcgs$dibpat, levels = c("Type A", "Type B"))
wcgs$chd69  <- factor(wcgs$chd69,  levels = c("No", "Yes"))
```

# Learning Objectives {.unnumbered}

By the end of this chapter, you should be able to:

1.  State the situation in which a **single sample mean** is compared with a **fixed (hypothesised) value**, and translate a research question into $H_0$/$H_1$.
2.  State the assumptions, formula, and correct interpretation of the **one-sample** $t$-test.
3.  State when a $z$-test is appropriate instead of a $t$-test, and carry out both by hand.
4.  Calculate the one-sample $t$-statistic and its confidence interval **by hand** from raw data.
5.  State the assumptions, formula, and correct interpretation of the **Wilcoxon signed-rank test**, the nonparametric alternative to the one-sample $t$-test.
6.  Calculate the Wilcoxon signed-rank statistic **by hand**, including handling of tied and zero differences.
7.  Reproduce all three procedures in **R**, using the Western Collaborative Group Study (**WCGS**) data set, and correctly choose among them for a given variable.

------------------------------------------------------------------------

# Comparing One Sample Mean to a Fixed Value

A very common question in health research is whether a group's average on some measurement differs from a known or hypothesised **standard**: a clinical guideline, a historical population value, a regulatory limit, or a target used in quality improvement. For example:

-   Is the mean systolic blood pressure of a cohort different from the widely used clinical benchmark of $120$ mmHg?
-   Does the mean height of a cohort of men differ from a commonly cited population reference of $69.0$ inches?
-   Is a cohort's mean total cholesterol above the "desirable" clinical threshold of $200$ mg/dL?

In each case there is **one sample** and **one fixed comparison value**, denoted $\mu_0$. The hypotheses are

$$
H_0: \mu = \mu_0 \qquad \text{vs.} \qquad H_1: \mu \ne \mu_0 \quad (\text{two-sided})
$$

or a one-sided alternative ($H_1: \mu > \mu_0$ or $\mu < \mu_0$) when justified by the research question.

Three procedures are commonly used, in decreasing order of how much they assume about the data:

| Test | Assumes population SD $\sigma$ known? | Assumes (approx.) normal data? | Uses |
|------------------|------------------|------------------|------------------|
| One-sample $z$-test | Yes | Yes | Raw values |
| One-sample $t$-test | No (estimates $\sigma$ with $s$) | Yes | Raw values |
| **Wilcoxon signed-rank test** | No | No — only symmetry around the centre | Ranks of the differences $x_i - \mu_0$ |

------------------------------------------------------------------------

# One-Sample $t$-Test (and $z$-Test When $\sigma$ Is Known) {#sec-t_z_test}

## Hypotheses

$$
H_0: \mu = \mu_0 \qquad H_1: \mu \ne \mu_0
$$

## Assumptions

1.  The outcome is **continuous (interval/ratio) numeric**.
2.  Observations are **independent** and represent a **random sample** from the population of interest.
3.  The population is (approximately) **normally distributed** — or $n$ is large enough that the Central Limit Theorem makes the sampling distribution of $\bar x$ approximately normal.
4.  For the $z$-test: the population standard deviation $\sigma$ is **known** (e.g., from an established reference population), not estimated from the sample.
5.  For the $t$-test (far more common in practice): $\sigma$ is **unknown** and estimated from the sample by $s$.

## Test statistics

$t$-test (the standard case: $\sigma$ unknown):

$$
t = \frac{\bar x - \mu_0}{s/\sqrt n} \qquad \sim \; t_{(n-1)} \text{ under } H_0
$$ {#eq-ttest}

$z$-test (only when $\sigma$ is genuinely known independently of the sample):

$$
z = \frac{\bar x - \mu_0}{\sigma/\sqrt n} \qquad \sim \; N(0,1) \text{ under } H_0
$$ {#eq-ztest}

In practice $\sigma$ is almost never known in advance, so the $t$-test is the default choice; the $z$-test is mostly a teaching device and is occasionally justified when a very well-established reference standard deviation is substituted for $\sigma$. Note that as $n$ grows large, $s \to \sigma$ and the $t$-distribution converges to the standard normal, so the two tests give nearly identical results in large samples (as we will see with the full WCGS data).

## Confidence interval for $\mu$

$$
\bar x \pm t_{(1-\alpha/2,\; n-1)} \cdot \frac{s}{\sqrt n} \qquad \text{(t-test version)}
$$

$$
\bar x \pm z_{(1-\alpha/2)} \cdot \frac{\sigma}{\sqrt n} \qquad \text{(z-test version)}
$$

## Interpretation

A significant result ($p < \alpha$, or a CI excluding $\mu_0$) indicates the sample mean is unlikely to have arisen from a population with mean $\mu_0$. As always, statistical significance should be reported **alongside** the estimated mean, the CI, and the magnitude of the difference — which carries the substantive/clinical meaning.

------------------------------------------------------------------------

# Manual (Hand) Calculation of the One-Sample $t$-Test

We reuse the same reproducible random subsample of $n=10$ men from the WCGS data used in the correlation chapter (`set.seed(123)`), and test whether their mean **height** differs from a fixed reference value of $\mu_0 = 69.0$ inches.

```{r}
#| label: t-manual-sample
#| echo: false
set.seed(123)
idx <- sample(seq_len(nrow(wcgs)), 10)
sub <- as.data.frame(wcgs[idx, c("id", "height")])
kable(sub, caption = "Random subsample (n = 10) used for the manual one-sample t-test")
```

## Step 1: Compute the sample mean and standard deviation

```{r}
#| label: t-manual-stats
#| echo: false
x    <- sub$height
n    <- length(x)
mu0  <- 69.0
xbar <- mean(x)
s    <- sd(x)
se   <- s / sqrt(n)

data.frame(n = n, mean = xbar, sd = s, se = se)
```

$$
\bar x = \frac{\sum x_i}{n} = \frac{694}{10} = 69.4, \qquad
s = \sqrt{\frac{\sum(x_i-\bar x)^2}{n-1}} = \sqrt{\frac{72.4}{9}} = 2.836
$$

$$
SE(\bar x) = \frac{s}{\sqrt n} = \frac{2.836}{\sqrt{10}} = 0.897
$$

## Step 2: Compute the test statistic

$$
t = \frac{\bar x-\mu_0}{SE(\bar x)} = \frac{69.4-69.0}{0.897} = \frac{0.4}{0.897}
$$

```{r}
#| label: t-manual-tstat
#| echo: false
t_stat <- (xbar - mu0) / se
df     <- n - 1
p_val  <- 2 * pt(-abs(t_stat), df = df)
data.frame(t = t_stat, df = df, p_value = p_val)
```

With $t = 0.446$ on $9$ df, $p = 0.666$: we **fail to reject** $H_0$ — this small sample provides no evidence that mean height differs from $69.0$ inches.

## Step 3: 95% confidence interval

$$
\bar x \pm t_{(0.975,\,9)}\times SE(\bar x) = 69.4 \pm 2.262 \times 0.897
$$

```{r}
#| label: t-manual-ci
#| echo: false
tcrit <- qt(0.975, df = df)
ci    <- xbar + c(-1, 1) * tcrit * se
data.frame(t_critical = tcrit, lower = ci[1], upper = ci[2])
```

## Step 4: Verify against R's built-in function

```{r}
#| label: t-manual-verify
t.test(x, mu = mu0)
```

The hand-calculated $\bar x$, $t$, $df$, $p$-value, and $95\%$ CI match R's `t.test()` output exactly.

------------------------------------------------------------------------

# Manual Demonstration of the $z$-Test ($\sigma$ Assumed Known)

Suppose, hypothetically, that a large historical reference population establishes the standard deviation of adult male height as $\sigma = 2.5$ inches, and we are willing to treat this as **known** rather than estimated from our own small sample. The $z$-test then uses $\sigma$ in place of $s$:

$$
z = \frac{\bar x-\mu_0}{\sigma/\sqrt n} = \frac{69.4 - 69.0}{2.5/\sqrt{10}}
$$

```{r}
#| label: z-manual
sigma0 <- 2.5
z_stat <- (xbar - mu0) / (sigma0 / sqrt(n))
p_z    <- 2 * pnorm(-abs(z_stat))
ci_z   <- xbar + c(-1, 1) * 1.96 * (sigma0 / sqrt(n))
data.frame(z = z_stat, p_value = p_z, ci_lower = ci_z[1], ci_upper = ci_z[2])
```

$z = 0.506$, $p = 0.613$ — the same conclusion as the $t$-test (no evidence against $H_0$), and the numeric value of $z$ is close to $t$ because $s=2.836$ and the assumed $\sigma=2.5$ are similar. Notice the $z$-test's CI ($67.85$ to $70.95$) is slightly **narrower** than the $t$-test's CI ($67.37$ to $71.43$) for the same data — using the standard normal critical value ($1.96$) instead of the larger $t$ critical value ($2.262$ at $9$ df) reflects the (arguably optimistic) assumption that $\sigma$ is known exactly rather than estimated with additional uncertainty from a small sample.

------------------------------------------------------------------------

# Wilcoxon Signed-Rank Test {#sec-wilcoxon}

## When to use it instead of the $t$-test

The Wilcoxon signed-rank test is the nonparametric counterpart to the one-sample $t$-test. Prefer it when:

-   The outcome is **ordinal**, or is numeric but **clearly non-normal / skewed** (as is common for laboratory values such as cholesterol, triglycerides, or cell counts).
-   The sample size is **small**, so the Central Limit Theorem cannot be relied upon to rescue non-normality.
-   There are **outliers** that would unduly influence the mean and $t$-statistic.

Rather than assuming normality, the Wilcoxon signed-rank test assumes only that the distribution of $x_i - \mu_0$ is **symmetric** about its centre; it tests whether that centre (formally, the median of the differences) equals zero.

$$
H_0: \text{median}(x_i - \mu_0) = 0 \qquad H_1: \text{median}(x_i-\mu_0)\ne 0
$$

## Procedure

1.  Compute the differences $d_i = x_i - \mu_0$ for each subject.
2.  **Discard** any $d_i = 0$ (subjects exactly equal to $\mu_0$), reducing $n$ accordingly.
3.  Rank the **absolute** differences $|d_i|$ from smallest (rank 1) to largest, giving **tied** values the average of the ranks they span.
4.  Attach back the original sign of $d_i$ to each rank ("signed ranks").
5.  Sum the ranks of the positive differences ($W^+$) and, separately, of the negative differences ($W^-$); note $W^+ + W^- = \dfrac{n(n+1)}{2}$ always.
6.  The test statistic is (by one common textbook convention) $W = \min(W^+, W^-)$; small values of $W$ are evidence against $H_0$. Software such as R instead commonly reports $V = W^+$ directly — both are exactly equivalent for testing purposes, since $W^+$ and $W^-$ carry the same information.
7.  For small $n$, compare $W$ to tables of exact critical values. For larger $n$ (a common rule of thumb: $n>20$, though software typically applies it more generally), use the normal approximation:

$$
z = \frac{W-\mu_W}{\sigma_W}, \qquad
\mu_W=\frac{n(n+1)}{4}, \qquad
\sigma_W=\sqrt{\frac{n(n+1)(2n+1)}{24}}
$$ {#eq-wilcoxon-z}

(A small downward correction to $\sigma_W$ is applied when there are tied ranks, and a continuity correction of $0.5$ is often applied to the numerator — both refinements are what statistical software does by default.)

## Assumptions and properties

-   Requires the outcome to be at least **ordinal**, and the distribution of differences to be (approximately) **symmetric** about its centre — a much weaker requirement than normality.
-   **Robust to outliers**, since only ranks (not raw magnitudes) enter the calculation.
-   Less **statistical power** than the $t$-test when the normality assumption genuinely holds, since it discards some numeric information by ranking.
-   Naturally estimates the **median** (or "pseudomedian" — the Hodges–Lehmann estimator) rather than the mean.

------------------------------------------------------------------------

# Manual (Hand) Calculation of the Wilcoxon Signed-Rank Test

We use the same reproducible random subsample of $n=10$ men (with complete cholesterol) from the correlation chapter (`set.seed(321)`), and test whether median **total cholesterol** differs from the clinical "desirable" threshold of $\mu_0=200$ mg/dL — a variable we already know is right-skewed, making this a good candidate for a nonparametric test.

```{r}
#| label: wilcoxon-manual-sample
#| echo: false
set.seed(321)
idx2 <- sample(which(!is.na(wcgs$chol)), 10)
sub2 <- as.data.frame(wcgs[idx2, c("id", "chol")])
kable(sub2, caption = "Random subsample (n = 10, complete cholesterol) used for the manual Wilcoxon signed-rank test")
```

## Step 1: Compute differences, rank absolute differences, reattach sign

Note subjects 3 and 10 are **tied** at $|d| = 5$ (chol $=205$, $d=5$): both receive the average rank $(1+2)/2 = 1.5$.

```{r}
#| label: wilcoxon-manual-table
#| echo: false
cx  <- sub2$chol
mu0c <- 200
d    <- cx - mu0c
absd <- abs(d)
rk   <- rank(absd)
signed_rk <- ifelse(d > 0, rk, -rk)

tab <- data.frame(id = sub2$id, chol = cx, d = d, abs_d = absd,
                   rank = rk, signed_rank = signed_rk)
kable(tab, col.names = c("ID","Cholesterol","$d=x-200$","$\\lvert d\\rvert$","Rank of $\\lvert d\\rvert$","Signed rank"))
```

## Step 2: Sum the positive and negative signed ranks

```{r}
#| label: wilcoxon-manual-sums
#| echo: false
Wpos <- sum(rk[d > 0])
Wneg <- sum(rk[d < 0])
n2   <- length(cx)
data.frame(`W+` = Wpos, `W-` = Wneg, `sum (check)` = Wpos + Wneg,
           `expected n(n+1)/2` = n2 * (n2 + 1) / 2, check.names = FALSE)
```

$W^+ = 44$, $W^- = 11$; the two sums correctly add to $n(n+1)/2 = 55$. Using the min-statistic convention, $W = \min(44,11) = 11$.

## Step 3: Normal approximation (since $n$ is small, this is illustrative; R uses an exact/tie-adjusted method by default)

$$
\mu_W = \frac{n(n+1)}{4} = \frac{10\times11}{4}=27.5,
\qquad
\sigma_W = \sqrt{\frac{n(n+1)(2n+1)}{24}} = \sqrt{\frac{10\times11\times21}{24}} = \sqrt{96.25}=9.811
$$

$$
z = \frac{W-\mu_W}{\sigma_W} = \frac{11-27.5}{9.811}
$$

```{r}
#| label: wilcoxon-manual-z
#| echo: false
mu_W    <- n2 * (n2 + 1) / 4
sigma_W <- sqrt(n2 * (n2 + 1) * (2 * n2 + 1) / 24)
W       <- min(Wpos, Wneg)
z_manual <- (W - mu_W) / sigma_W
p_manual <- 2 * pnorm(-abs(z_manual))
data.frame(mu_W, sigma_W, W, z = z_manual, p_value = p_manual)
```

By hand (uncorrected normal approximation): $z=-1.68$, two-sided $p \approx 0.093$ — a borderline result, **not** significant at the conventional $5\%$ level for this small subsample.

## Step 4: Verify against R's built-in function

```{r}
#| label: wilcoxon-manual-verify
#| echo: false
wilcox.test(cx, mu = mu0c)                       # default: continuity-corrected normal approx (ties present)
wilcox.test(cx, mu = mu0c, correct = FALSE)       # without continuity correction
```

R reports $V = 44$ — this is simply $W^+$, the sum of the positive signed ranks, which is R's chosen statistic (equivalent information to our $W=\min(W^+,W^-)=11$). With the continuity correction, $p=0.103$; without it, $p=0.092$, matching our hand calculation closely (the tiny remaining difference is R's additional correction to $\sigma_W$ for the tied rank at $1.5$). Both the manual and software results reach the same conclusion: no significant evidence against $H_0$ in this small subsample.

------------------------------------------------------------------------

# Demonstration in R Using the Full WCGS Data Set

## Variable 1: Height vs. a fixed reference of 69.0 inches (approximately normal - good case for $t$/$z$)

```{r}
#| label: fig-height-hist
#| fig-cap: "Distribution of height in the WCGS cohort (n = 3,154), with the hypothesised value marked."
ggplot(wcgs, aes(x = height)) +
  geom_histogram(binwidth = 1, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = 69.0, colour = "firebrick", linewidth = 1, linetype = "dashed") +
  labs(x = "Height (inches)", y = "Count",
       title = "Height distribution vs. hypothesised mean (dashed line, μ₀ = 69.0 in)") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: full-t-height
t.test(wcgs$height, mu = 69.0)
```

**Interpretation:** mean height $= 69.78$ in, $95\%$ CI $(69.69,\,69.87)$, $t(3153)=17.27$, $p<0.001$. With the full cohort, even the small $0.78$-inch difference from $69.0$ is highly statistically significant, because the huge sample size gives enormous precision — again a reminder to interpret the **magnitude** of the difference (well under one inch), not only the $p$-value.

```{r}
#| label: full-z-height
xbar_full <- mean(wcgs$height); n_full <- length(wcgs$height)
z_full <- (xbar_full - 69.0) / (2.5 / sqrt(n_full))
p_full <- 2 * pnorm(-abs(z_full))
data.frame(mean = xbar_full, n = n_full, z = z_full, p_value = p_full)
```

The $z$-test (assuming $\sigma=2.5$ known) gives $z=17.47$, $p\approx2\times10^{-68}$ — the same conclusion as the $t$-test. With $n=3{,}154$, $s=2.53$ is so close to the assumed $\sigma=2.5$, and the $t$-distribution with $3{,}153$ df is so close to standard normal, that the two procedures are practically indistinguishable — illustrating why the $t$-test is used almost universally in practice: it gives the (safer) $t$-test answer even when a large sample would make a $z$-test numerically almost identical.

## Variable 2: Cholesterol vs. a fixed clinical threshold of 200 mg/dL (right-skewed - good case for Wilcoxon)

```{r}
#| label: fig-chol-hist
#| fig-cap: "Distribution of total cholesterol in the WCGS cohort, showing the right skew and the hypothesised threshold."
ggplot(wcgs, aes(x = chol)) +
  geom_histogram(binwidth = 10, fill = "darkorange3", colour = "white") +
  geom_vline(xintercept = 200, colour = "firebrick", linewidth = 1, linetype = "dashed") +
  labs(x = "Total cholesterol (mg/dL)", y = "Count",
       title = "Cholesterol distribution vs. hypothesised value (dashed line, μ₀ = 200 mg/dL)") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: full-normcheck-chol
shapiro.test(wcgs$chol)
```

As previously established, cholesterol is significantly right-skewed (Shapiro-Wilk $p<0.001$; recall this test is oversensitive at $n>3{,}000$, but the histogram confirms genuine skew) — a textbook case for preferring a nonparametric approach, or at least reporting it alongside the $t$-test.

```{r}
#| label: full-t-chol
t.test(wcgs$chol, mu = 200)
```

```{r}
#| label: full-wilcoxon-chol
wilcox.test(wcgs$chol, mu = 200, conf.int = TRUE)
```

**Interpretation:** the $t$-test gives mean $=226.4$ mg/dL, $95\%$ CI $(224.9,\,227.9)$, $t(3141)=34.0$, $p<0.001$. The Wilcoxon test gives an estimated (pseudo)median of $225$ mg/dL, $95\%$ CI $(223.5,\,226.5)$, $p<0.001$. Both tests agree emphatically that cholesterol in this cohort is higher than the $200$ mg/dL threshold — the effect here is so large that the choice of test does not change the conclusion, but the **Wilcoxon estimate (pseudomedian** $225$) is closer to the true sample median ($223$) than the mean ($226.4$), which is pulled upward by the right skew — illustrating why the median/pseudomedian can be the more representative summary for skewed clinical measurements.

```{r}
#| label: full-median-mean
data.frame(mean_chol = mean(wcgs$chol, na.rm = TRUE),
           median_chol = median(wcgs$chol, na.rm = TRUE))
```

## Side-by-side summary across several variables

```{r}
#| label: tbl-summary-onesample
vars_list <- list(
  list(var = "height", mu0 = 69.0),
  list(var = "bmi",    mu0 = 25.0),
  list(var = "sbp",    mu0 = 120.0),
  list(var = "chol",   mu0 = 200.0)
)

results <- do.call(rbind, lapply(vars_list, function(v) {
  x  <- wcgs[[v$var]]
  ok <- !is.na(x)
  tt <- t.test(x[ok], mu = v$mu0)
  wt <- wilcox.test(x[ok], mu = v$mu0)
  data.frame(Variable = v$var, mu0 = v$mu0, n = sum(ok),
             Sample_mean = round(mean(x[ok]), 2),
             Sample_median = round(median(x[ok]), 2),
             t_stat = round(unname(tt$statistic), 2),
             t_p = signif(tt$p.value, 3),
             Wilcoxon_V = unname(wt$statistic),
             Wilcoxon_p = signif(wt$p.value, 3))
}))
kable(results)
```

Systolic blood pressure (`sbp`) vs. $120$ mmHg and BMI (`bmi`) vs. $25.0$ kg/m$^2$ are included here as further examples — in both cases the cohort mean is significantly above the reference value, consistent with the WCGS cohort being composed of middle-aged, largely overweight men in the 1960s, and both parametric and nonparametric tests agree in direction and significance despite the skew present in these variables.

------------------------------------------------------------------------

# Choosing Among the Three Tests

| Situation | Recommended test |
|------------------------------------|------------------------------------|
| $\sigma$ genuinely known from an external, trusted source | $z$-test |
| $\sigma$ unknown (the usual case), data approximately normal or $n$ large | One-sample $t$-test |
| Data ordinal, or numeric but clearly skewed / outlier-prone, especially with small $n$ | Wilcoxon signed-rank test |
| Data extremely skewed with meaningful zeros/ties, or very small $n$ with clear non-normality | Wilcoxon signed-rank test (or consider an exact sign test) |

------------------------------------------------------------------------

# Common Pitfalls

-   **Using a** $z$-test when $\sigma$ is not truly known. Substituting the sample SD $s$ into the $z$-formula and calling it a "$z$-test" understates uncertainty (the $t$-distribution's heavier tails exist precisely to correct for estimating $\sigma$) — always use the $t$-test unless $\sigma$ comes from an independent, trusted source.
-   **Ignoring skewness.** A $t$-test is fairly robust to non-normality with large $n$ (as the height and cholesterol examples show, both parametric and nonparametric tests agreed here), but with small samples and marked skew, results can diverge — always plot the distribution first.
-   **Confusing statistical and clinical/practical significance.** With $n>3{,}000$, trivial differences from $\mu_0$ become "significant"; always report and interpret the estimated effect size (mean or median difference and CI), not just the $p$-value.
-   **Forgetting that the Wilcoxon test targets the median (or centre of symmetry), not the mean.** When comparing results between a $t$-test and a Wilcoxon test, remember they are (strictly) testing different population parameters, which will usually agree in direction but need not be numerically identical.
-   **Mishandling ties and zero differences in the Wilcoxon test.** Zero differences ($x_i=\mu_0$) must be dropped and $n$ reduced accordingly; tied absolute differences must receive averaged ranks, and software applies a variance correction for these ties that a naive hand calculation may omit.
-   **Non-independent or non-random samples.** Both tests assume the observations are an independent random sample from the population of interest; a convenience sample or clustered/repeated data will invalidate the standard errors and $p$-values of either test.

------------------------------------------------------------------------

# Summary

-   To compare a single sample mean to a fixed value $\mu_0$, use the **one-sample** $t$-test by default (it is valid under normality or, via the CLT, for large $n$).
-   Use the $z$-test only when $\sigma$ is genuinely known independently of the sample; in practice this is rare, and as $n$ grows large the $t$-test and $z$-test converge, so there is little cost to always using the $t$-test.
-   Use the **Wilcoxon signed-rank test** when the outcome is ordinal, markedly skewed, or contains outliers, particularly with small samples — it tests the median/centre of symmetry using signed ranks rather than raw values, and is robust to non-normality.
-   In the WCGS data, mean height ($69.78$ in) is (statistically, if not clinically) significantly different from $69.0$ inches by both $t$- and $z$-tests, and mean/median cholesterol ($226.4$ / $223$ mg/dL) is significantly above the $200$ mg/dL clinical threshold by both the $t$-test and the Wilcoxon test — with the Wilcoxon's pseudomedian estimate ($225$) illustrating how it better represents a right-skewed distribution than the mean does.

------------------------------------------------------------------------

# References {.unnumbered}

-   Rosner B. *Fundamentals of Biostatistics*. 8th ed. Cengage Learning.
-   Daniel WW, Cross CL. *Biostatistics: A Foundation for Analysis in the Health Sciences*. 11th ed. Wiley.
-   Bland M. *An Introduction to Medical Statistics*. 4th ed. Oxford University Press.
-   Woodward M. *Epidemiology: Study Design and Data Analysis*. 3rd ed. Chapman & Hall/CRC.
-   Wilcoxon F. "Individual Comparisons by Ranking Methods." *Biometrics Bulletin*. 1945;1(6):80-83.
-   Rosenman RH, et al. "A predictive study of coronary heart disease: the Western Collaborative Group Study." *JAMA*. 1964;189:15-22. (Original WCGS study.)
