---
title: "Comparing Two Paired (Matched) Measurements"
output:
  pdf_document:
    number_sections: yes
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_knit$set(root.dir = "D:/Epi Biostat with R/Biostat_with_R/")
```

```{r}
#| label: setup
#| include: false
library(readxl)
library(ggplot2)
library(knitr)
library(car)

options(digits = 4)
wcgs <- readxl::read_excel("Section2/Categorical data analysis/Datasets/wcgs.xls")
wcgs$dibpat <- factor(wcgs$dibpat, levels = c("Type A", "Type B"))
wcgs$chd69  <- factor(wcgs$chd69,  levels = c("No", "Yes"))
```

# Learning Objectives {.unnumbered}

By the end of this chapter, you should be able to:

1.  Recognise a **paired (matched) design** — where two measurements are linked because they come from the *same* subject (or a matched pair) — and explain why this is different from comparing two independent groups.
2.  State the hypotheses, assumptions, formula, and correct interpretation of the **paired** $t$-test, and explain why it is mathematically identical to a one-sample $t$-test performed on the differences.
3.  Calculate the paired $t$-test **by hand**, including the deviation table used to obtain the standard deviation of the differences.
4.  State the assumptions, formula, and correct interpretation of the **Wilcoxon signed-rank test** as a paired/matched-samples procedure.
5.  Calculate the Wilcoxon signed-rank statistic **by hand** for paired data, including handling of tied ranks.
6.  Reproduce all procedures in **R**, using the WCGS data, and correctly explain why ignoring pairing (treating matched data as if it were two independent samples) produces the wrong standard error.

------------------------------------------------------------------------

# Comparing Two Paired (Matched) Measurements {#sec-paired_t}

A **paired** (or **matched**) design arises whenever two measurements are linked at the level of the individual observational unit — most commonly because the *same subject* contributes both values. Classic examples are "before vs. after treatment" on the same patient, or two raters scoring the same specimen. The defining statistical feature is not the timing of the measurements but the fact that the two values are **correlated**: knowing one tells you something about the other, because both come from the same source of between-subject variability.

The WCGS data set does not contain a longitudinal before/after design, but it does contain something equally legitimate for a paired analysis: **systolic blood pressure (`sbp`)** and **diastolic blood pressure (`dbp`)**, both measured on the *same man at the same clinic visit*. These two numbers are far from independent — a man with an unusually high `sbp` for his age and build will tend to also have an unusually high `dbp`, because both partly reflect the same underlying vascular physiology. That within-subject correlation is precisely the situation the paired test is built to exploit, and it is worth confirming formally before proceeding:

```{r}
#| label: sbp-dbp-cor
cor.test(wcgs$sbp, wcgs$dbp)
```

The correlation is strong and highly significant ($r = 0.773$, $95\%$ CI $0.758$ to $0.787$, $p<0.001$). Because `sbp` and `dbp` are this strongly linked within a person, a test that **ignores** the pairing (i.e., treats the two columns as if they came from two different, independent groups of men) throws away real information and — as we will demonstrate later in this chapter — produces the *wrong* standard error for the comparison.

The paired difference itself is not merely a technical device here; it is a genuine, clinically meaningful quantity:

$$
d_i = \text{sbp}_i - \text{dbp}_i = \textbf{pulse pressure}
$$

Pulse pressure is a recognised cardiovascular measure related to arterial stiffness. Because systolic pressure is, by definition of the cardiac cycle, essentially always higher than diastolic pressure for a given person, we already know with near certainty that mean pulse pressure will be positive and that a paired test will reject $H_0:\mu_d=0$ — that direction is a matter of physiology, not a "finding." The genuinely useful statistical content in this chapter is therefore twofold: (a) demonstrating paired-test technique on real, honestly-paired data, and (b) estimating the **size and precision** of mean pulse pressure in this cohort — a legitimate epidemiological quantity — while showing exactly how the paired approach correctly accounts for the sbp/dbp correlation that an independent two-sample test would ignore.

```{r}
#| label: fig-sbp-dbp-scatter
#| fig-cap: "Systolic vs. diastolic blood pressure, same subject, same visit — the visible positive association is exactly why these two measurements should be analysed as a paired sample."
ggplot(wcgs, aes(x = dbp, y = sbp)) +
  geom_point(alpha = 0.15, colour = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick") +
  labs(x = "Diastolic BP (mmHg)", y = "Systolic BP (mmHg)") +
  theme_minimal(base_size = 12)
```

Two procedures are commonly used for paired data:

| Test | Assumes normality? | Uses |
|----|----|----|
| **Paired** $t$-test | Yes, of the *differences* $d_i$ (or large $n$) | Raw differences |
| **Wilcoxon signed-rank test** | No — only (approximate) symmetry of the differences | Ranks of $\lvert d_i\rvert$ |

------------------------------------------------------------------------

# The Paired $t$-Test {#sec-paired_t_test}

## Hypotheses

$$
H_0: \mu_d = 0 \qquad H_1: \mu_d \ne 0
$$

where $\mu_d$ is the population mean of the paired differences $d_i = x_{1i} - x_{2i}$.

## Assumptions

1.  The design is genuinely **paired/matched**: each $d_i$ comes from a single subject (or matched pair) contributing both measurements.
2.  The differences $d_i$ are **independent across subjects** (subject $j$'s difference tells us nothing about subject $k$'s) — independence is required *between* pairs, not within a pair (within a pair, correlation is expected and is exactly what pairing accounts for).
3.  The differences $d_i$ are **(approximately) normally distributed**, or $n$ is large enough for the Central Limit Theorem to apply to $\bar d$. Critically, it is the distribution of the *differences* that must be checked — **not** the distributions of the two raw variables, which may each be non-normal even when their differences are well behaved (or vice versa).

## Test statistic

$$
t = \frac{\bar d}{s_d/\sqrt n} \qquad \sim\; t_{(n-1)} \text{ under } H_0
$$ {#eq-pairedt}

where $\bar d = \frac{1}{n}\sum d_i$ and $s_d = \sqrt{\dfrac{\sum (d_i-\bar d)^2}{n-1}}$ is the standard deviation of the differences.

## Confidence interval for $\mu_d$

$$
\bar d \pm t_{(1-\alpha/2,\;n-1)} \cdot \frac{s_d}{\sqrt n}
$$

## Connection to the one-sample $t$-test

Compare @eq-pairedt with the one-sample $t$-statistic from [previous chapter](#sec-t_z_test), $t=(\bar x-\mu_0)/(s/\sqrt n)$. They are exactly the same formula: the paired $t$-test **is** a one-sample $t$-test applied to the single derived variable $d_i$, tested against the fixed value $\mu_0=0$. There is no new machinery here — only a new variable to feed into the familiar one-sample procedure. This is why, in R, `t.test(x1, x2, paired = TRUE)` and `t.test(x1 - x2, mu = 0)` give identical output, as we verify below.

------------------------------------------------------------------------

# Manual (Hand) Calculation of the Paired $t$-Test

We draw a reproducible random subsample of $n=10$ men (`set.seed(444)`) and compute their pulse pressure $d = \text{sbp}-\text{dbp}$.

```{r}
#| label: paired-manual-sample
#| echo: false
set.seed(444)
sub <- as.data.frame(wcgs[sample(nrow(wcgs), 10), c("id", "sbp", "dbp")])
sub$d <- sub$sbp - sub$dbp
kable(sub, caption = "Random subsample (n = 10) with paired sbp, dbp, and difference d = sbp - dbp (pulse pressure)")
```

## Step 1: Build the deviation table

```{r}
#| label: paired-manual-dev
#| echo: false
n    <- nrow(sub)
dbar <- mean(sub$d)
dev  <- sub$d - dbar
dev2 <- dev^2
 
tab <- data.frame(id = sub$id, sbp = sub$sbp, dbp = sub$dbp, d = sub$d,
                   dev = dev, dev2 = dev2)
kable(tab, digits = 2,
      col.names = c("ID", "SBP", "DBP", "d", "d - dbar", "(d - dbar)^2"),
      caption = "Deviation table for the paired t-test")
```

$$
\bar d = \frac{\sum d_i}{n} = \frac{484}{10} = 48.4
$$

```{r}
#| label: paired-manual-sum-dev2
#| echo: false
sum_dev2 <- sum(dev2)
sum_dev2
```

## Step 2: Standard deviation and standard error of the differences

$$
s_d = \sqrt{\frac{\sum (d_i-\bar d)^2}{n-1}} = \sqrt{\frac{590.4}{9}} = 8.099
$$

```{r}
#| label: paired-manual-sd
#| echo: false
sd_d <- sqrt(sum_dev2 / (n - 1))
se_d <- sd_d / sqrt(n)
data.frame(n = n, mean_d = dbar, sd_d = sd_d, se_d = se_d)
```

## Step 3: Test statistic and $p$-value

$$
t = \frac{\bar d}{s_d/\sqrt n} = \frac{48.4}{2.561}
$$

```{r}
#| label: paired-manual-t
t_stat <- dbar / se_d
df_d   <- n - 1
p_val  <- 2 * pt(-abs(t_stat), df = df_d)
data.frame(t = t_stat, df = df_d, p_value = p_val)
```

## Step 4: 95% confidence interval

```{r}
#| label: paired-manual-ci
tcrit <- qt(0.975, df = df_d)
ci    <- dbar + c(-1, 1) * tcrit * se_d
data.frame(t_critical = tcrit, lower = ci[1], upper = ci[2])
```

## Step 5: Verify against R

```{r}
#| label: paired-manual-verify
t.test(sub$sbp, sub$dbp, paired = TRUE)
t.test(sub$d, mu = 0)     # identical, confirming the one-sample-on-differences equivalence
```

The hand-calculated $\bar d = 48.4$, $s_d = 8.099$, $t = 18.90$, $df = 9$, $p = 1.50\times 10^{-8}$ match R's `t.test(sub$sbp, sub$dbp, paired = TRUE)` exactly, and `t.test(sub$d, mu = 0)` reproduces the identical output — confirming the equivalence stated above. Even in this tiny subsample, mean pulse pressure is enormous relative to its standard error (as physiology predicts), so the result is not remotely borderline.

------------------------------------------------------------------------

# The Wilcoxon Signed-Rank Test for Paired Data {#sec-new_testw}

## When to use it instead of the paired $t$-test

The Wilcoxon signed-rank test is the nonparametric counterpart to the paired $t$-test (it is the *same* procedure introduced in [previous chapter](#sec-paired_t) for one-sample data — applied here to the paired differences $d_i = x_{1i}-x_{2i}$ instead of $x_i-\mu_0$). Prefer it when:

-   The differences $d_i$ are **clearly non-normal / skewed**, or contain influential **outliers**.
-   The outcome is **ordinal** rather than a proper interval scale.
-   The sample size is **small**, so the Central Limit Theorem cannot be relied upon to rescue non-normal differences.

It requires only that the distribution of the differences be **(approximately) symmetric** about its centre, a much weaker assumption than normality:

$$
H_0: \text{median}(d_i) = 0 \qquad H_1: \text{median}(d_i) \ne 0
$$

## Procedure

1.  Compute the differences $d_i = x_{1i} - x_{2i}$ for each pair.
2.  **Discard** any $d_i = 0$ (exact ties between the pair), reducing $n$ accordingly.
3.  Rank the **absolute** differences $\lvert d_i \rvert$ from smallest (rank 1) to largest, giving **tied** absolute differences the average of the ranks they span.
4.  Attach back the original sign of $d_i$ to each rank.
5.  Sum the ranks of the positive differences ($W^+$) and, separately, of the negative differences ($W^-$); note $W^+ + W^- = \dfrac{n(n+1)}{2}$ always.
6.  The test statistic is (by one common textbook convention) $W=\min(W^+,W^-)$; small values of $W$ are evidence against $H_0$. R instead reports $V=W^+$ directly — equivalent information.
7.  For larger $n$, use the normal approximation:

$$
z = \frac{W-\mu_W}{\sigma_W}, \qquad
\mu_W=\frac{n(n+1)}{4}, \qquad
\sigma_W=\sqrt{\frac{n(n+1)(2n+1)}{24}}
$$ {#eq-wilcoxon-paired-z}

(software applies a downward correction to $\sigma_W$ for tied ranks, and often a $0.5$ continuity correction to the numerator).

------------------------------------------------------------------------

# Manual (Hand) Calculation of the Wilcoxon Signed-Rank Test

We reuse the **same** $n=10$ paired subsample from the previous section. Notice from the deviation table that two pairs share $\lvert d\rvert = 44$ (subjects with $d=44,44$) and two share $\lvert d\rvert = 46$ ($d=46,46$) — genuine ties in the absolute differences, giving a useful worked example of tie handling. Also notice that **every** $d_i$ is positive: as discussed above, this is expected on physiological grounds (sbp exceeds dbp for essentially every subject) and is not itself informative — the interesting content is in the ranks' magnitudes, not their signs.

## Step 1: Rank the absolute differences and reattach sign

```{r}
#| label: wilcoxon-paired-manual-table
#| echo: false
absd <- abs(sub$d)
rk   <- rank(absd)
signed_rk <- ifelse(sub$d > 0, rk, -rk)
 
tabw <- data.frame(id = sub$id, d = sub$d, abs_d = absd, rank = rk, signed_rank = signed_rk)
kable(tabw, col.names = c("ID", "d", "$\\lvert d\\rvert$", "Rank of $\\lvert d\\rvert$", "Signed rank"),
      caption = "Ranking of absolute paired differences (n = 10)")
```

## Step 2: Sum the positive and negative signed ranks

```{r}
#| label: wilcoxon-paired-manual-sums
#| echo: false
Wpos <- sum(rk[sub$d > 0])
Wneg <- sum(rk[sub$d < 0])
data.frame(`W+` = Wpos, `W-` = Wneg, `sum (check)` = Wpos + Wneg,
           `expected n(n+1)/2` = n * (n + 1) / 2, check.names = FALSE)
```

$W^+=55$, $W^-=0$ — every difference is positive, so all of the rank mass falls on the positive side; the two sums correctly add to $n(n+1)/2=55$. Using the min-statistic convention, $W=\min(55,0)=0$.

## Step 3: Normal approximation

$$
\mu_W = \frac{n(n+1)}{4} = \frac{10\times 11}{4} = 27.5,
\qquad
\sigma_W = \sqrt{\frac{n(n+1)(2n+1)}{24}} = \sqrt{\frac{10\times 11\times 21}{24}} = \sqrt{96.25} = 9.811
$$

$$
z = \frac{W-\mu_W}{\sigma_W} = \frac{0-27.5}{9.811}
$$

```{r}
#| label: wilcoxon-paired-manual-z
#| echo: false
mu_W    <- n * (n + 1) / 4
sigma_W <- sqrt(n * (n + 1) * (2 * n + 1) / 24)
W       <- min(Wpos, Wneg)
z_manual <- (W - mu_W) / sigma_W
p_manual <- 2 * pnorm(-abs(z_manual))
data.frame(mu_W, sigma_W, W, z = z_manual, p_value = p_manual)
```

By hand (uncorrected normal approximation): $z=-2.803$, two-sided $p\approx 0.0051$.

## Step 4: Verify against R

```{r}
#| label: wilcoxon-paired-manual-verify
wilcox.test(sub$sbp, sub$dbp, paired = TRUE)                  # default: continuity-corrected
wilcox.test(sub$sbp, sub$dbp, paired = TRUE, correct = FALSE) # matches the hand normal approximation
```

R reports $V=55$ — exactly our $W^+$. Without the continuity correction, $p=0.00500$, matching our hand calculation ($p\approx0.0051$) closely (the tiny remaining gap is R's additional variance correction for the tied ranks at $3.5$ and $5.5$); with the continuity correction (R's default), $p=0.00586$. Both are far below $0.05$, and both agree with the paired $t$-test: pulse pressure is significantly different from zero even in this $n=10$ subsample — again, an expected physiological result, with the genuinely interesting information residing in the *size* of the effect rather than its direction.

------------------------------------------------------------------------

# Demonstration in R Using the Full WCGS Data Set

## Distribution of pulse pressure and checking the actual assumption (normality of the differences)

The assumption that matters for the paired $t$-test is normality of $d_i = \text{sbp}_i-\text{dbp}_i$ — **not** normality of `sbp` or `dbp` individually.

```{r}
#| label: fig-pulse-pressure
#| fig-cap: "Distribution of pulse pressure (sbp - dbp) in the full WCGS cohort (n = 3,154)."
#| fig-subcap:
#|   - "Histogram of pulse pressure"
#|   - "Normal Q-Q plot of pulse pressure"
#| layout-ncol: 2
wcgs$pulse_pressure <- wcgs$sbp - wcgs$dbp
 
ggplot(wcgs, aes(x = pulse_pressure)) +
  geom_histogram(binwidth = 2, fill = "steelblue", colour = "white") +
  labs(x = "Pulse pressure = SBP - DBP (mmHg)", y = "Count") +
  theme_minimal(base_size = 12)
 
qqnorm(wcgs$pulse_pressure, main = "")
qqline(wcgs$pulse_pressure, col = "firebrick", lwd = 2)
```

```{r}
#| label: full-shapiro-pp
shapiro.test(wcgs$pulse_pressure)
```

The histogram is unimodal and reasonably symmetric with a slight right skew, and the Q-Q plot shows only modest departure from the reference line in the tails. Shapiro-Wilk nonetheless rejects exact normality ($p<0.001$) — but with $n=3{,}154$, Shapiro-Wilk is well known to detect even trivially small deviations from normality as "significant" (a large-$n$ oversensitivity we have flagged in previous chapters). Combined with $n$ being very large (so the Central Limit Theorem applies comfortably to $\bar d$ regardless of the shape of individual differences) and the mild skew visible in the plots, the paired $t$-test remains entirely appropriate here; we run the Wilcoxon test alongside it as a robustness check.

## Paired $t$-test on the full sample

```{r}
#| label: full-paired-t
t.test(wcgs$sbp, wcgs$dbp, paired = TRUE)
```

Mean pulse pressure is $46.62$ mmHg ($95\%$ CI $46.28$ to $46.96$), $t(3153)=267.4$, $p<0.001$. With $n=3{,}154$, this huge $t$-statistic is not remarkable in itself (the direction was guaranteed); what is useful is the tight confidence interval — we can state mean pulse pressure in this cohort to within about $\pm0.34$ mmHg.

## Wilcoxon signed-rank test on the full sample

```{r}
#| label: full-wilcoxon-pp
wilcox.test(wcgs$sbp, wcgs$dbp, paired = TRUE, conf.int = TRUE)
```

The Wilcoxon test gives an estimated (pseudo)median difference (the Hodges-Lehmann estimator) of $46.0$ mmHg, $95\%$ CI $(46.00, 46.00)$ — reported to more decimal places by R as $45.99997$ to $46.00004$, i.e. an extremely tight interval given the huge $n$ — and $p<0.001$. The $t$-test mean ($46.62$) and the Wilcoxon pseudomedian ($46.0$) are close but not identical, consistent with the mild right skew seen in the histogram; both procedures agree emphatically on the substantive conclusion.

## What goes wrong if pairing is ignored

Suppose, mistakenly, that `sbp` and `dbp` were treated as two independent samples rather than paired measurements on the same men (i.e., an (incorrect) unpaired/Welch two-sample test):

```{r}
#| label: full-unpaired-wrong
t.test(wcgs$sbp, wcgs$dbp, paired = FALSE, var.equal = FALSE)
```

```{r}
#| label: compare-se
se_paired   <- unname((t.test(wcgs$sbp, wcgs$dbp, paired = TRUE)$conf.int[2] -
                        t.test(wcgs$sbp, wcgs$dbp, paired = TRUE)$estimate) / qt(0.975, 3153))
se_unpaired <- unname(sd(wcgs$sbp)^2/length(wcgs$sbp) + sd(wcgs$dbp)^2/length(wcgs$dbp))
se_unpaired <- sqrt(se_unpaired)
data.frame(design = c("Paired (correct)", "Independent (incorrect)"),
           SE = c(se_paired, se_unpaired))
```

The point estimate of the mean difference is unchanged ($46.62$ mmHg either way, since both are just $\bar{\text{sbp}}-\bar{\text{dbp}}$), but the standard error nearly **doubles** when the (incorrect) independent-samples approach is used ($SE\approx0.320$ vs. the correct paired $SE\approx0.174$), and correspondingly the (incorrect) $t$-statistic drops from $267.4$ to $145.6$ and the confidence interval widens. This is exactly the cost of throwing away the strong positive within-subject correlation ($r=0.773$) confirmed at the start of the chapter: the independent-samples formula assumes zero correlation between the two sets of values and so overstates the combined uncertainty. In general, whenever two measurements are legitimately paired, using an independent-samples test does not just lose power — it computes the wrong standard error for the design that was actually used to collect the data.

## Comparison table

```{r}
#| label: tbl-summary-paired
tt <- t.test(wcgs$sbp, wcgs$dbp, paired = TRUE)
wt <- wilcox.test(wcgs$sbp, wcgs$dbp, paired = TRUE, conf.int = TRUE)
 
results <- data.frame(
  Test = c("Paired t-test", "Wilcoxon signed-rank"),
  Estimate = c(unname(tt$estimate), unname(wt$estimate)),
  CI_lower = c(tt$conf.int[1], wt$conf.int[1]),
  CI_upper = c(tt$conf.int[2], wt$conf.int[2]),
  Statistic = c(unname(tt$statistic), unname(wt$statistic)),
  df_or_NA = c(unname(tt$parameter), NA),
  p_value = c(signif(tt$p.value, 3), signif(wt$p.value, 3))
)
kable(results, digits = 3,
      col.names = c("Test", "Estimate (mean or pseudomedian)", "95% CI lower", "95% CI upper",
                    "Statistic (t or V)", "df", "p-value"),
      caption = "Paired t-test vs. Wilcoxon signed-rank test for pulse pressure (sbp - dbp), full WCGS cohort")
```

------------------------------------------------------------------------

# A Practical Decision Process

1.  Confirm the design is genuinely **paired/matched** (same subject or matched unit contributes both values) — if not, use the independent-groups methods of the previous chapter instead.
2.  Compute the differences $d_i$ and **check their distribution** (histogram, Q-Q plot; Shapiro-Wilk interpreted cautiously in large samples) — not the distributions of the two raw variables.
3.  Choose accordingly:

| Differences approximately normal (or large $n$)? | Recommended test |
|----|----|
| Yes | Paired $t$-test |
| No (skewed, outlier-prone, ordinal, or small $n$ with clear non-normality) | Wilcoxon signed-rank test |

4.  Report the **estimate and confidence interval** (mean difference or pseudomedian), not just the $p$-value — especially important in a paired design where a large $n$ can make even a small, physiologically uninteresting difference "significant."

------------------------------------------------------------------------

# Common Pitfalls

-   **Mistaking a paired design for an independent one.** As shown above, running an (incorrect) independent two-sample test on data that is actually paired ignores the within-subject correlation and produces the wrong (here, nearly doubled) standard error — it does not just cost power, it computes the wrong quantity for the design used.
-   **Checking normality of the raw variables instead of the differences.** The paired $t$-test's normality assumption applies to $d_i=x_{1i}-x_{2i}$, not to `sbp` or `dbp` individually; a variable can be skewed on its own while its paired differences with another correlated variable are well behaved, or vice versa.
-   **Forgetting to drop or correctly handle zero differences and ties in the Wilcoxon test.** Exact zero differences ($d_i=0$) must be removed and $n$ reduced accordingly; tied $\lvert d_i\rvert$ values must receive averaged ranks, and software applies a variance correction for ties that a naive hand calculation may omit.
-   **Over-interpreting a physiologically (or logically) guaranteed direction as a "finding."** Because SBP essentially always exceeds DBP within a person, obtaining $p<0.001$ for $H_0:\mu_d=0$ tells us nothing we didn't already know — the informative output of this analysis is the **size and precision** of mean pulse pressure ($46.6$ mmHg, tight CI), not its sign.
-   **Confusing statistical and clinical/physiological significance.** With $n=3{,}154$, the confidence interval for mean pulse pressure is extremely tight; always report and interpret the magnitude and CI, not only whether $p<0.05$.

------------------------------------------------------------------------

# Summary

-   A **paired (matched) design** links two measurements at the level of the individual subject; the WCGS `sbp` and `dbp` values, measured on the same man at the same visit, are strongly correlated ($r=0.773$) and constitute a legitimate paired sample, even though the study was not a before/after design.
-   The **paired** $t$-test tests $H_0:\mu_d=0$ on the differences $d_i=x_{1i}-x_{2i}$ and is mathematically identical to a one-sample $t$-test applied to those differences ($t=\bar d/(s_d/\sqrt n)$); on our $n=10$ hand-calculation subsample, $t=18.90$, $df=9$, $p=1.50\times10^{-8}$, matching R exactly.
-   The **Wilcoxon signed-rank test** is the nonparametric counterpart, requiring only symmetry of the differences; in the same subsample, $W^+=55$, $W^-=0$, normal-approximation $z=-2.80$, $p\approx0.0051$, closely matching R's uncorrected result ($p=0.00500$).
-   On the **full cohort** ($n=3{,}154$), mean pulse pressure is $46.62$ mmHg ($95\%$ CI $46.28$ to $46.96$; Wilcoxon pseudomedian $46.0$ mmHg) — a precise, clinically interpretable estimate.
-   Treating `sbp` and `dbp` as **independent** rather than paired nearly doubles the standard error ($0.320$ vs. $0.174$) for the identical point estimate, illustrating concretely why pairing must be respected in the analysis whenever it is present in the data-collection design.

------------------------------------------------------------------------

# References {.unnumbered}

-   Rosner B. *Fundamentals of Biostatistics*. 8th ed. Cengage Learning.
-   Daniel WW, Cross CL. *Biostatistics: A Foundation for Analysis in the Health Sciences*. 11th ed. Wiley.
-   Bland M. *An Introduction to Medical Statistics*. 4th ed. Oxford University Press.
-   Wilcoxon F. "Individual Comparisons by Ranking Methods." *Biometrics Bulletin*. 1945;1(6):80-83.
-   Rosenman RH, et al. "A predictive study of coronary heart disease: the Western Collaborative Group Study." *JAMA*. 1964;189:15-22. (Original WCGS study.)
